Prophage-encoded Hm-oscar gene recapitulates Wolbachia-induced male killing in the tea tortrix moth Homona magnanima

Version 1

Abstract

Abstract

Wolbachia are the maternally transmitted bacterial symbionts that are ubiquitous among arthropods. They can hijack host reproduction in various ways, including male killing (MK), where the sons of infected mothers are killed during development. The recent discovery of MK-associated Wolbachia genes, i.e., oscar in Ostrinia moths and wmk in Drosophila flies, stimulates our interest in the diversity and commonality of MK mechanisms, which remain largely unclear. We recently discovered that a Wolbachia symbiont of the moth Homona magnanima carries an MK-associated prophage region encoding homologs of oscar (Hm-oscar) and wmk (wmk-1–4). Here, we investigated the effects of these genes in the native host. Upon transient overexpression, Hm-oscar, but not wmk, induced male lethality in H. magnanima, in contrast to our observations in Drosophila, where the wmk homologs, but not Hm-oscar, killed the males. Hm-oscar disrupted sex determination in male embryos by inducing a female-type doublesex splicing and impaired dosage compensation, recapitulating the Wolbachia phenotype. Cell-based transfection assays confirmed that Hm-oscar suppressed the function of masculinizer, the upstream male sex determinant involved in lepidopteran dosage compensation. Our study highlights the conserved roles of oscar homologs in Wolbachia-induced lepidopteran MK and argues that Wolbachia have evolved multiple MK mechanisms in insects.

Introduction

Arthropods commonly carry microbial symbionts that are passed from mother to offspring (Hurst & Frost, 2015; Werren et al., 2008; Hurst, 2017). The maternally transmitted bacteria, belonging to the genus Wolbachia (Alphaproteobacteria), are estimated to be present in at least 40% of all insect species, making them one of the most widespread endosymbiont genera on the planet (Werren et al, 2008; Zug & Hammerstein, 2012). Wolbachia have achieved evolutionary success by manipulating host reproduction through various means that enhance endosymbiont transmission (Werren et al., 2008). Such manipulation of the host’s reproduction includes cytoplasmic incompatibility (CI), parthenogenesis, feminization, and male killing (MK), each of which seemingly affects the biological features, distribution, and evolution of the host. Among these strategies, MK directly skews the sex ratio of the host population toward females by killing male offspring of infected mothers during development. The lack of symbiont transmission through male hosts often leads to the evolution of MK. Wolbachia have been shown to induce MK in a wide range of insect taxa (e.g., dipterans, lepidopterans, and coleopterans) and other arthropods (Hurst et al., 1999; Kageyama & Traut, 2004; Hurst et al., 2000). Furthermore, various bacteria, viruses, and microsporidia induce MK phenotypes (Kageyama et al., 2012; Fujita et al., 2021; Kageyama et al., 2023; Nagamine et al., 2023), and recent studies have postulated that these microbes have evolved their MK ability independently (Harumoto & Lemaitre, 2018; Kageyama et al., 2023; Nagamine et al., 2023; Arai et al., 2023a).

The evolution and molecular mechanisms underlying Wolbachia-induced MK have attracted considerable attention for decades. It has been hypothesized that the mechanisms underlying MK are associated with sex determination cascades in insects (Hornett et al., 2022). Indeed, some MK-inducing Wolbachia strains have been shown to directly interact with the sex determination system in lepidopteran insects (Sugimoto and Ishikawa, 2012; Sugimoto et al., 2015; Fukui et al., 2015; Arai et al., 2023a). The Lepidopteran sex determination system consists of multiple transcriptional regulators, some of which exhibit sex-linked expression and/or splicing isoforms. Lepidoptera generally have a female heterogametic sex chromosome system (e.g., WZ/ZZ) and employ dosage compensation, which equalizes the sex-linked (Z-chromosome-linked) gene dose between males and females (Kiuchi et al., 2014). Dosage compensation is regulated by the masculinizer gene (masc), which is critical for male development. masc also controls the downstream master regulator of sex determination and differentiation, doublesex (dsx), which exhibits sex-dependent splicing isoforms (dsxF in females and dsxM in males) (Kiuchi et al., 2014). In Ostrinia and Homona male moths, MK-inducing Wolbachia strains impair dosage compensation by disrupting the expression of masc. Furthermore, they disrupt sex determination in male moths by inducing the “female” isoform of dsx (dsxF), leading to a “mismatch” between genetic sex (male: ZZ sex chromosome constitution) and phenotypic sex (female: based on dsxF) and consequently to male death (Arai et al., 2023a; Fukui et al., 2015; Sugimoto et al., 2015; Sugimoto and Ishikawa, 2012).

More recently, a protein produced in Wolbachia wFur strain named Oscar (Osugoroshi protein containing CifB C-terminus-like domain and many Ankyrin Repeats; Osugoroshi translates to male killing in Japanese) was shown to recapitulate the Wolbachia-induced MK phenotype in Ostrinia (Katsuma et al., 2022). The Oscar protein degrades and interacts with Masc protein, leading to the failure of dosage compensation and the production of female-type dsx isoforms in Ostrinia male moths (Fukui et al., 2024; Katsuma et al., 2022). Although Oscar homologs have been identified in several MK-inducing Wolbachia strains in Lepidoptera, some MK Wolbachia strains in Diptera (Drosophila) and Lepidoptera do not carry this protein (Arai et al., 2023b; 2024a). Furthermore, Oscar homologs are evolutionarily dynamic, with highly variable sequences and structures (classified as Type I and Type II Oscar) (Arai et al., 2024a), making it difficult to assess their functional relevance. In addition to Oscar, the helix-turn-helix domain-containing putative transcriptional regulator Wmk (approximately 300 aa) is widely conserved among Wolbachia strains and induces various toxicities in Drosophila flies (i.e., no effect, weak male lethality [30% of males die], strong male lethality [90%–100% of males die], and death of all males and females) (Perlmutter et al., 2019, 2020, 2021; Arai et al., 2023b). Although the mechanisms underlying the Wmk-induced toxicities and their connection to sex determination systems remain unclear, these findings suggest that Wolbachia strains carry multiple factors that cause male lethality. However, the diversity and commonality of these functions in insects remain largely unknown, partly due to the technical challenges in validating gene functions in non-model insects.

We recently discovered an MK-associated prophage region that underlies the evolutionary transition from a non-MK Wolbachia (wHm-c) to an MK Wolbachia (wHm-t) in the tea tortrix moth Homona magnanima (Tortricidae) (Arai et al., 2023b; Arai et al., 2024b). The MK-associated prophage element encodes four wmk homologs (wmk-1 to wmk-4) as well as an oscar homolog (Hm-Oscar, 1181 aa), which differs significantly in sequence length and structure from the wFur-encoded Oscar (1830 aa). In Drosophila, Wmk-1 and Wmk-3 have a lethality of almost 100%, while Wmk-2, Wmk-4, and Hm-Oscar induce no lethal effects when singly overexpressed. Furthermore, co-expression of the adjacent Wmk-3 and Wmk-4 has been shown to induce the death of 90% of male flies and restores female survival (Arai et al., 2023b), suggesting that their combined action is similar to that observed in Wolbachia-induced CI (CifA and CifB, Beckmann et al., 2017; LePage et al., 2017). However, the mechanistic links between prophage-encoded Wolbachia genes and MK in the native host Homona remain unclear.

In this study, we showed that the prophage-encoded Hm-oscar recapitulates Wolbachia-induced MK in H. magnanima. Furthermore, we clarified the mechanistic links to host sex-determination cascades both in vivo and in vitro and discussed the underlying mechanisms of MK in Lepidoptera, arguing for the diverse evolutionary origin of Wolbachia-induced MK.

Results

Hm-Oscar induces female-biased sex ratios

To achieve the transient overexpression of Hm-oscar and the four wmk genes (wmk-1, wmk-2, wmk-3, and wmk-4), constructed mRNA (cRNA) was injected into Wolbachia-free H. magnanima embryos. Subsequently, the adult moths that emerged from the cRNA-injected embryos were sexed based on their external morphology. When Hm-oscar was overexpressed, the sex ratio of adults was strongly female-biased (85.6% ± 11.1%; 42 males and 218 females in total, 15 replicates), which was in sharp contrast with the ratios observed in the GFP-injected (50.9% ± 2.08%; 126 females and 123 males in total, 10 replicates) and non-injected (NSR) (48.6% ± 3.08%; 201 females and 215 males in total, 9 replicates) groups (P = 0.001 and 0.002, respectively, Steel–Dwass test, Fig. 1). Compared with the GFP-injected group, the sex ratio was not biased by the overexpression of wmk-1 (48.9% ± 7.27%, 6 replicates, P = 0.999), wmk-2 (46.8% ± 4.77%, 6 replicates, P = 0.810), wmk-3 (46.7% ± 7.91%, 5 replicates, P = 0.999), and wmk-4 (47.1% ± 3.47%, 5 replicates, P = 0.773). Although the dual expression of wmk-3 and wmk-4 induces strong male lethality in Drosophila (Arai et al., 2023b), here, it did not skew the sex ratio of Homona compared with the ratio detected in the GFP-injected group (42.1% ± 7.67%, 5 replicates, P = 0.704). Similarly, the dual expression of the tandemly arrayed wmk-1 and wmk-2 did not bias the sex ratio (50.6% ± 2.82%, 5 replicates) compared with the value in the GFP-injected group (P = 1.000).

Transient expressions of the Hm-oscar gene resulted in female-biased sex ratios

Male ratio of adult progeny obtained from cRNA-injected groups (wmk-1, wmk-2, T2A-brigged wmk-1 and wmk-2, wmk-3, wmk-4, T2A-brigged wmk-3 and wmk-4, GFP, and Hm-oscar; n = 5–15 independent replicates using different egg masses), the non-injected NSR line, and the MK wHm-t-positive WT12 line. The total numbers of adult females and males are shown at the bottom. Different letters indicate significant differences (Steel–Dwass test, P < 0.05). The dot plots show all data points individually.

Males are killed mainly during the embryonic stage

The sex of unhatched embryos and hatched larvae (neonates) was determined by karyotyping for W chromatin, with the presence and absence of this substance indicating females and males, respectively. In the Hm-oscar-injected group, the sex ratio of the hatched larvae (neonates) was strongly female-biased (21 females and 3 males, P = 0.0001 in the binomial test) (Fig. 2a). In contrast, that of unhatched mature embryos was male-biased (18 females and 38 males, P = 0.005), suggesting that males died mainly during the embryonic stage. Likewise, the sex ratios of hatched larvae (34 females and 6 males) and unhatched embryos (27 females and 47 males) of the MK wHm-t infected line (WT12) were also female- and male-biased, respectively (P = 0.013 and 0, respectively). In the non-injected group (NSR), the sex ratios of hatched larvae (28 females and 28 males) and unhatched embryos (23 females and 25 males) were not biased (P = 0.553 and 0.443, respectively).

Hm-Oscar induced lethality in male embryos with the female-type sex determination

(a) Sex ratio of the hatched larvae and unhatched embryos in the Hm-oscar-expressed, wHm-t-infected, and non-infected/expressed groups. Females and males were discriminated based on the presence or absence of W chromatin. The number of individuals is indicated in brackets. (b) Splicing patterns of the downstream sex-determining gene dsx of H. magnanima embryos (5 days post oviposition). Abbreviations: HmOsc, Hm-oscar injected group; GFP, GFP-injected group; WT12, wHm-t-infected line; NSR, non-infected/injected line. M and F indicate W chromatin-negative (ZZ: male genotype) and W chromatin-positive (ZW: female genotype) mature embryos, respectively. dsx-F and dsx-M represent female and male-specific splicing variants, respectively.

Female-type sex determination in male embryos that are destined to die

Splicing patterns of the downstream sex determinant dsx were assessed in each of the H. magnanima embryos that were Hm-oscar-expressed, GFP-expressed, wHm-t-infected, or non-expressed (i.e., non-injected, NSR). In GFP-expressed and NSR embryos, females and males (which were identified by the presence and absence of W chromatin, respectively) exhibited a female- and male-type splicing variant of dsx, respectively. However, both wHm-t-infected and Hm-oscar-expressed embryos induced female-type dsx splicing regardless of the presence or absence of W chromatin (Fig. 2b).

Hm-Oscar impairs dosage compensation in male embryos

RNA-seq analysis revealed that, in Hm-oscar-injected embryos, Z-linked genes (homologs on the Bombyx mori chromosomes 1 and 15) were more expressed in males than in females (Fig. 3a), which was not observed in the GFP-injected group (Fig. 3b). Similarly, as previously reported by Arai et al. (20-23a), high levels of Z-linked gene expression were also observed in wHm-t-infected males, but not in NSR males (Figure 3c,d).

Hm-oscar-overexpressed male embryos showed higher levels of Z-linked gene expression

(a-d) Normalized expression levels (TPM) and chromosomal distributions of transcripts in H. magnanima embryos. RNA-seq data of embryos (108 hpo) were used to make the following comparisons: Hm-oscar-injected males versus Hm-oscar injected females (a), GFP-injected males versus GFP-injected females (b), WT12 males versus WT12 females (c), and NSR males versus NSR females (d). The chromosome number for each H. magnanima transcript-derived contig was assigned based on Bombyx mori gene models. The X axis represents the chromosome number of B. mori (shown as chr01 to chr28), and chr01 and chr15 (highlighted in orange) correspond to the Z chromosome of H. magnanima (Arai et al., 2023a).

Hm-Oscar suppresses the masculinizing functions of lepidopteran Masc

To confirm whether Hm-Oscar suppresses the functions of the upstream male sex determinant Masc, their interactions were assessed by transfection using BmN-4 cells, as described in Katsuma et al. (2022). BmN-4 cells, which are derived from the female ovaries of B. mori (Grace, 1967), exhibit a female-type default sex determination (Kiuchi et al., 2014). In contrast with the control groups, which were transfected with the non-inserted pIZ/V5-His vector and showed this default sex determination, the cells transfected with the Hm-masc-inserted pIZ/V5-His vector exhibited the male-type sex determination, as evidenced by the increased expression levels of the male-specific splicing of the sex-determining gene BmImpM (Fig. 4). Furthermore, the masculinizing function of Hm-masc was suppressed by co-transfecting the Hm-masc/Hm-oscar-inserted pIZ/V5-His vectors, as manifested by the low expression levels of BmImpM. While type I oscar of the MK wFur strain also suppressed Hm-masc, the function of this gene was less active than that of Hm-oscar. Hm-oscar also suppressed the masculinizing function of masc genes derived from various lepidopteran insects (i.e., Ostrinia furnacalis, Spodoptera frugiperda, Bombyx mori, Trichoplusia ni, and Papilio machaon), suggesting its broad spectrum of action in Lepidoptera.

Hm-Oscar suppressed the masculinizing function of lepidopteran Masc

Relative expression levels of the male-specific BmImpM variant in transfected BmN-4 cells. The expression of BmImpM was normalized using the housekeeping gene rp49. The expression levels of BmImpM in the Masc and Hm-Oscar/Oscar co-transfected cells were normalized by setting each Masc-transfected cell as 100. The dot plots show all data points individually. Each experimental condition was replicated three times. Abbreviations: Hm, Homona magnanima; Bm, Bombyx mori; Of, Ostrinia furnacalis; Pm, Papilio machaon; Sf, Spodoptra frugiperda; and Tv, Trilocha varians.

Discussion

In this study, we showed that the phage-encoded Hm-oscar, but not wmk, induced male lethality and a female-biased sex ratio in H. magnanima. Furthermore, the overexpressed Hm-oscar impaired male sex determination in Homona, recapitulating the wHm-t-induced phenotypes. Cell-based assays confirmed that Hm-Oscar suppressed the masculinizing functions of Masc. These results strongly suggested that Hm-Oscar underlies the wHm-t-induced MK function in H. magnanima. The Wolbachia-encoded Oscar homologs identified so far are classified as type I and type II (Arai et al., 2024a). Although long ankyrin repeat sequences are postulated to be critical for the function of wFur-encoded Oscar (1830 aa, type I) (Katsuma et al., 2022a), our study revealed that the functions of Hm-Oscar (1181 aa, type II), which encodes fewer ankyrin repeats, were comparable with those of Oscar carried by wFur. Interestingly, type II Oscar is also present in the feminizing Wolbachia wFem in butterflies belonging to genus Eurema (Arai et al., 2024a). Oscar homologs, which inhibit the masculinizing function and induce female sex determination, may have a conserved function in Wolbachia-induced MK and feminization in Lepidoptera.

In contrast to the results of this study, we have previously demonstrated that the phage-encoded wmk, but not Hm-oscar, induces male lethality in D. melanogaster (Arai et al., 2023b). Although the means of expression are different (i.e., transient in Homona and transgenic in Drosophila), this finding highlighted the differences in the mode/range of action of Wolbachia genes between insect species. It has been hypothesized that microbes induce MK in insects by targeting molecular mechanisms involved in sex determination and differentiation (Hornett et al., 2022b). Sex determination systems in insects are diverse; for example, Lepidoptera (including H. magnanima) and Diptera (including D. melanogaster) do not share any known sex-determining genes other than dsx (Suzuki, 2018). The different outcomes in Homona and Drosophila are probably due to their different sex determination systems. Because Oscar interacts with and suppresses Masc, Hm-Oscar could induce mortality in Homona males that possess Masc, but not in Drosophila males that lack it. Considering that Oscar homologs are not present in known MK Wolbachia strains in dipteran insects (Arai et al., 2024c; Katsuma et al., 2022), the mechanisms (i.e., causative genes) of Wolbachia-induced MK probably differ between insect taxa (e.g., between Lepidoptera and Diptera). While the mechanisms underlying Wmk-induced lethality remain unclear, the distinct effects associated with this gene between Homona and Drosophila may also reflect their genetic backgrounds [e.g., presence/absence of host factor(s) that interact with Wmk]. In addition, Katsuma et al. (2022) reported that the Wmk homologs encoded by wFur did not affect the masculinizing function of Masc in vitro, indicating that Wmk likely targets factors other than Masc. Our results strongly suggested that Wolbachia strains have acquired multiple MK genes through evolution. An intense evolutionary arms race between Wolbachia and their hosts may have increased the diversity of MK-inducing genes in the Wolbachia genome.

Wolbachia-induced phenotypes are known to be influenced by the genetic backgrounds of hosts (Hornett et al., 2006; Sasaki et al., 2002; Veneti et al., 2012). Our study showed that the wHm-t-encoded Hm-Oscar suppresses the function of Hm-Masc in H. magnanima more efficiently than the wFur-encoded Oscar, suggesting that this Wolbachia factor has undergone evolutionary tuning to adapt to its natural host. However, the mere presence of Oscar and Wmk homologs does not ensure the expression of the MK phenotype. For example, the type I Oscar-bearing wCauA strain did not induce MK in Ephestia (Cadra) cautella collected around 2000, although it did induce MK when transferred to the closely related host Ephestia kuehniella (Sasaki et al., 2002). However, an MK phenotype, presumably induced by wCauA, was recorded in E. cautella around the 1970s (Takahashi & Kuwahara, 1970). These findings suggested the emergence of suppressor(s) against MK induced by Oscar-bearing wCauA in E. cautella. Virulence genes often undergo duplication and substitution under strong selective pressure (Hill et al., 2022; Jones & Dangl, 2006). An intense evolutionary arms race between Wolbachia and their hosts could have increased and diversified the MK-associated Wolbachia genes. Conversely, natural selection favors the rescue of males by suppressing the Wolbachia-induced reproductive manipulations (Hornett et al., 2006, 2014), which may involve changes in the sex determination system because Wolbachia strains frequently hijack host reproduction systems. In this context, Wolbachia and other MK-inducing microbes may be a hidden driver for the diversification of complex insect sex determination systems.

In this study, we clarified the conserved roles of the Wolbachia-encoded Oscar homologs in Lepidoptera and demonstrated that Wolbachia have evolved distinct MK mechanisms (through causative genes) in insect taxa. The diversification of phenotype/virulence-associated genes and the rampant horizontal transmission of phages carrying virulence genes between Wolbachia strains may have contributed to the outstanding success of this bacterial genus. In addition to oscar and wmk, Wolbachia may retain other uncharacterized genes that induce male lethality, and further studies on diverse Wolbachia–host systems are highly warranted. Our findings provide insights into the molecular mechanisms and evolutionary relationships between endosymbionts and their hosts, which may also contribute to the design of pest management strategies.

Materials and methods

Experimental model and subject details

Homona magnanima

A laboratory-maintained H. magnanima line with a normal sex ratio (NSR) that was negative for Wolbachia and other endosymbionts, was used in our experiments. This line was initially collected in Tokyo, Japan, in 1999 and has been maintained inbred for over 250 generations in the laboratory. Larvae were reared using the artificial SilkMate 2S diet (Nosan Co., Yokohama, Japan) at 25°C under a long photoperiod (16L:8D) (Arai et al., 2019).

The laboratory-maintained all female H. magnanima WT12 line, which was initially collected in Taiwan (Tea Research Extension Station, Taoyuan city) in 2015 (Arai et al., 2020), was also used in this study. This line was maintained for over 50 generations by crossing it with the males of the NSR line.

BmN-4 cell line

BmN-4 cells (provided by Chisa Yasunaga-Aoki, Kyushu University, and maintained in our laboratory) were cultured at 26°C in IPL-41 Insect Medium (Applichem, Darmstadt, Germany) supplemented with 10% fetal bovine serum.

Transient expression of MK-associated phage genes

(i) mRNA synthesis

Codon-optimized wmk genes (wmk-1 to wmk-4), conjugated wmk pairs using a T2A peptide (i.e., wmk-1-T2A-wmk-2 and wmk-3-T2A-wmk-4), and Hm-oscar genes synthesized by Arai et al. (2023b) were used for mRNA synthesis. These synthetic genes were ligated into the plasmid pIZ/V5-His (Invitrogen, MA, USA) using the NEBuilder® HiFi DNA Assembly kit (New England Biolabs, MA, USA) following the manufacturer’s protocol. The inserts of the vector (i.e., wmk-1, wmk-2, wmk-3, wmk-4, Hm-oscar, wmk-1-T2A-wmk-2, wmk-3-T2A-wmk-4, and GFP as a control gene) were amplified using KOD-one (TOYOBO, Osaka, Japan) with the primer set containing the T7 promoter described in Fukui et al. (2015) (i.e., pIZ-F-T7: 5′-TAATACGACTCACTATAGGGAGACAGTTGAACAGCATCTGTTC-3′ and pIZ-R: 5′-GACAATACAAACTAAGATTTAGTCAG-3′) under the following PCR conditions: 20 cycles at 98°C for 10 s, 62°C for 5 s, and 68°C for 15 s. The amplicons were purified using the QIAquick PCR purification kit (Qiagen, Hilden, Germany), and 100–200 ng DNA was used for mRNA synthesis. The capped mRNA (cRNA) with poly(A) tail was synthesized using the mMESSAGE mMACHINE® T7 Ultra kit (Invitrogen, MA, USA) following the manufacturer’s protocol with some modifications. In brief, the assembled transcription reaction (10 µL of T7 2X NTP/ARCA, 2 µL of 10X T7 reaction buffer, 2 µL of T7 enzyme mix, and 6 µL of linear DNA template diluted with nuclease-free water) was incubated at 37°C for at least 10 h to maximize RNA yields. After poly(A) tailing, cRNA was purified using ISOGEN II (Nippongene, Tokyo, Japan) and dissolved in up to 10 µL of nuclease-free water to achieve an RNA concentration of approximately 1500–2000 ng/µL. The synthesized cRNA was preserved at −80°C until further use.

(ii) Preparation of H. magnanima eggs

A total of 15 males and 10 females of H. magnanima were mated in a plastic box (30 cm × 20 cm × 5 cm) for 3–4 days. The collection of egg masses began the day after oviposition was confirmed. In brief, newly oviposited egg masses were collected at 30-min intervals during the dark period using red light (which the moths cannot perceive). Females started to oviposit eggs at least 4 h into the dark period and, within less than 30 min post oviposition (mpo), the egg masses were collected. The collection lasted until the start of the light period. The egg masses were then subjected to microinjection assays.

(iii) Inoculum preparation, microinjection, and maintenance of the injected embryos

A glass needle for microinjection was prepared using glass capillary GD-1 (Narishige, Tokyo, Japan) with a PC-10 puller (Narishige). The glass capillaries were pulled at two temperatures (first stage: 75°C, second stage: 65°C) using two heavy weights and one light weight. The movement position during the second heating stage was set to 3 mm (range 1–10 mm).

The cRNA solution was diluted in a buffer (100 mM potassium acetate, 2 mM magnesium acetate, and 30 mM HEPES-KOH; at pH 7.4) containing 0.2% (W/V) Brilliant Blue FCF (Wako, Osaka, Japan) to obtain an RNA concentration of 1000 ng/µL. Approximately 1–4 µL of dye-containing mRNA solution was aspirated into the glass needles (capillaries), and the edge of each needle was ground using Micro Grinder EG-402 (Narishige) at an angle of 20 degrees for 2 s.

The fresh egg masses of H. magnanima (collected at less than 30 mpo as described above) were put on a double-sided sticky tape (15 mm × 5 m T4612, Nitoms, Tokyo, Japan) and placed on glass slides. Under a Nikon SMZ1270 microscope (Nicon, Tokyo, Japan), the RNA solution (30–100 fL) was injected into individual eggs using an microinjector IM-400 (Narishige) (balance pressure set at 0.010–0.030; injection pressure set at 0.050-0.100) and a QP-3JOY-2R electric micromanipulator (MicroSupport, Shizuoka, Shizuoka).

The injected egg masses were maintained in a 90-mm plastic Petri dish fitted with a slightly wet filter paper. As high humidity interfered with the development of the embryos, the injected eggs (egg mass) were first maintained at a relative humidity of 30%–50% (0–3 days post injection) and then at a higher humidity until hatching (4–6 dpo, 60% relative humidity). The hatched larvae were reared separately with 1/2 ounces of SilkMate 2S (Nosan Co.) until eclosion. The adult H. magnanima moths that emerged from the cRNA-injected embryos were sexed based on their external morphology.

Sexing of embryos/neonates and RNA extraction

To verify the effect of the transient expression of Hm-oscar on sex determination in H. magnanima, Hm-oscar/GFP-expressed, non-injected, and wHm-t-infected mature embryos showing black head capsule (1 day before hatching, 5–6 days post oviposition, dpo) were dissected on glass slides using forceps, as described in Arai et al. (2022). Their Malpighian tubules were fixed with methanol/acetic acid (50% v/v) and stained with lactic acetic orcein for W chromatin observations. The remaining tissues not used for sexing were stored in ISOGEN II (Nippon Gene) at −80°C until subsequent extraction. In total, 12 males or females (confirmed based on the presence or absence of W chromatin) were pooled and homogenized in ISOGEN II to extract RNA as described in Arai et al. (2023a). In brief, 0.4 times the volume of UltraPure distilled water (Invitrogen) was added to the ISOGEN II homogenates, which were centrifuged at 12000 × g for 15 min at 4°C to pellet proteins and DNAs. The resulting supernatant was mixed with the same volume of isopropanol to precipitate RNAs; then, the resulting solutions were transferred to EconoSpin columns (Epoch Life Science) and centrifuged at 17,900 × g and 4°C for 2 min. The RNAs captured in the column were washed twice with 80% ethanol and eluted in 20 µL of UltraPure distilled water (Invitrogen).

Hmdsx detection

Sex-specific dsx splicing variants of H. magnanima were detected as described in Arai et al. (2023a). In brief, total RNA (100–300 ng) extracted from sex-determined mature embryos was reverse-transcribed using PrimeScript™ II Reverse Transcriptase (TaKaRa, Shiga, Japan) at 30°C for 10 min, 45°C for 60 min, and 70°C for 15 min. Then, cDNA was amplified using KOD-FX Neo (Toyobo Co., Ltd.) with the following two primers: Hmdsx_long3F (5′-TGCCTAAAGTGAAAACGCCGAGGAGCC-3′) and Hmdsx_Mrev (5′-TGGAGGTCTCTTTTCATCCGG-3′). The PCR conditions used were as follows: 94°C for 2 min, followed by 45 cycles of 98°C for 10 s, 66°C for 30 s, and 68°C for 30 s. The amplicons were subjected to electrophoresis on 2.0% agarose Tris-borate-EDTA buffer (89 mM Tris-borate, 89 mM boric acid, 2 mM EDTA) gels.

RNA-seq-based quantification of Z chromosome-linked genes

The effects on dosage compensation were assessed by measuring differences in gene expression, as described in Fukui et al. (2015) and Arai et al. (2023a). A total of 1.0 µg of the total RNA extracted from Hm-oscar or GFP-overexpressed mature embryos (108 hpo) was used to prepare mRNA-seq libraries via the NEBNext Poly (A) mRNA Magnetic Isolation Module (New England Biolabs) and the NEBNext Ultra II RNA Library Prep kit for Illumina (New England Biolabs) following the manufacturer’s protocol. The adaptor sequences and low-quality reads (Qscore <20) were removed from the generated sequence data [150 bp paired-end (PE150)] using Trimmomatic (Bolger et al., 2014). The trimmed reads were mapped to the previously assembled transcriptome database for H. magnanima (Arai et al., 2023a) using Kallisto (Bray et al., 2016) to generate the normalized read count data (transcripts per million, TPM).

The binary logarithms of the TPM differences between males and females belonging to each H. magnanima group (i.e., Oscar/GFP-expressed) were calculated to assess the fold-changes in gene expression levels between sexes. As described in Arai et al. (2023a), the transcriptome data of H. magnanima were annotated using the B. mori gene sets obtained from KAIKOBASE (https://kaikobase.dna.affrc.go.jp). The binary logarithms of the TPM differences between males and females in the B. mori chromosomes 1 to 28 were plotted. The expression data of the wHm-t-infected and non-infected groups were also calculated based on the transcriptome data included in Arai et al. (2023a).

Transfection assays and quantification of BmIMP

The coding sequence of Hm-masc was amplified from cDNA derived from the RNA extracted from male embryos of the NSR line using KOD-one (TOYOBO) with the following primer set: HmMasc_CDS_HindIIIF: 5′-GCAAAGCTTCAACATGATCTCTCGCCAACCACAATCAACATCA-3′ and HmMasc_CDS_BamHI R: 5′-GCAGGATCCCAACCTACTGATAAGGAGGGAAGTAAGGCTGCTG-3′. The following

PCR conditions were applied: 20 cycles of 98°C for 10 s, 62°C for 5 s, and 68°C for 15 s. The amplicon was purified with the QIAquick PCR purification kit (Qiagen) and cloned into the pIZ/V5-His vector using the HindIII and BamHI enzymes (New England Biolabs). The codon-optimized oscar gene of wFur (Katsuma et al. 2022) was also cloned into the pIZ/V5-His vector using the KpnI and NotI enzymes (TaKaRa).

To verify the masculinizing function of Hm-masc, BmN-4 cells (4 × 105 cells per dish, diameter 35 mm) were transfected with 1 µg of plasmid DNA (pIZ/V5-His having Hm-masc) using FuGENE HD (Promega, WI, USA), as described in Katsuma et al. (2022). To clarify whether Hm-oscar suppressed the function of Hm-masc, 1 µg of plasmid DNA (pIZ/V5-His bearing either Hm-masc or Hm-oscar) was co-transfected to the BmN-4 cells using FuGENE HD (Promega). Three days after transfection, the cells were collected and subjected to RNA extraction via TRI REAGENT® (Molecular Research Center Inc., USA) and cDNA construction with AMV transcriptase (TaKaRa). The degree of masculinization in the BmN-4 cells (default female-type sex determination) was verified by quantifying the expression levels of BmImpM, which is involved in male-specific sex determination cascades, using primers rp49_F: 5′-CCCAACATTGGTTACGGTTC-3′ and rp49_R: 5′-GCTCTTTCCACGATCAGCTT-3′; BmIMP_F: 5′-ATGCGGGAAGAAGGTTTTATG-3′ and BmIMP_R: 5′-

TCATCCCGCCTCAGACGATTG-3′, as described in Fukui et al. (2015). Further interactions between Hm-Oscar and Masc proteins derived from lepidopteran insects [i.e., Trilocha varians Masc (TvMasc), Spodoptra frugiperda Masc (SfMasc), B. mori Masc (BmMasc), O. furnacalis Masc (OfMasc), and Papilio machaon Masc (PmMasc)] (Katsuma et al., 2022a) were assessed using the same procedures.

Quantification and statistical analysis

The number of surviving H. magnanima injected with either Hm-oscar, wmks, or GFP in each replicate were counted at the adult stage. The male ratios (number of adult males/numbers of all adults) under all conditions were compared using the Steel– Dwass test in R v4.0. The sex ratio bias was also assessed based on the total numbers of (i) male and female adults and (ii) male and female embryos under each condition using the binomial test in R v4.0.

eLife Assessment

This valuable study, which implicates a specific Wolbachia gene in driving the male-killing phenotype in a moth, contributes to a growing body of literature from the authors in which they have nicely teased apart the loci responsible for male killing across diverse insects. Solid evidence supports the conclusions, though improvements to the statistical analysis for certain assays would strengthen the inferences further.

Reviewer #1 (Public review):

Summary:

Insects and their relatives are commonly infected with microbes that are transmitted from mothers to their offspring. A number of these microbes have independently evolved the ability to kill the sons of infected females very early in their development; this male killing strategy has evolved because males are transmission dead-ends for the microbe. A major question in the field has been to identify the genes that cause male killing and to understand how they work. This has been especially challenging because most male-killing microbes cannot be genetically manipulated. This study focuses on a male-killing bacterium called Wolbachia. Different Wolbachia strains kill male embryos in beetles, flies, moths, and other arthropods. This is remarkable because how sex is determined differs widely in these hosts. Two Wolbachia genes have been previously implicated in male-killing by Wolbachia: oscar (in moth male-killing) and wmk (in fly male-killing). The genomes of some male-killing Wolbachia contain both of these genes, so it is a challenge to disentangle the two.

This paper provides strong evidence that oscar is responsible for male-killing in moths. Here, the authors study a strain of Wolbachia that kills males in a pest of tea, Homona magnanima. Overexpressing oscar, but not wmk, kills male moth embryos. This is because oscar interferes with masculinizer, the master gene that controls sex determination in moths and butterflies. Interfering with the masculinizer gene in this way leads the (male) embryo down a path of female development, which causes problems in regulating the expression of genes that are found on the sex chromosomes.

Strengths:

The authors use a broad number of approaches to implicate oscar, and to dissect its mechanism of male lethality. These approaches include:

(1) Overexpressing oscar (and wmk) by injecting RNA into moth eggs.

(2) Determining the sex of embryos by staining female sex chromosomes.

(3) Determining the consequences of oscar expression by assaying sex-specific splice variants of doublesex, a key sex determination gene, and by quantifying gene expression and dosage of sex chromosomes, using RNASeq.

(4) Expressing oscar along with masculinizer from various moth and butterfly species, in a silkmoth cell line.

This extends recently published studies implicating oscar in male-killing by Wolbachia in Ostrinia corn borer moths, although the Homona and Ostrinia oscar proteins are quite divergent. Combined with other studies, there is now broad support for oscar as the male-killing gene in moths and butterflies (i.e. order Lepidoptera). So an outstanding question is to understand the role of wmk. Is it the master male-killing gene in insects other than Lepidoptera and if so, how does it operate?

Weaknesses:

I found the transfection assays of oscar and masculinizer in the silkworm cell line (Figure 4) to be difficult to follow. There are also places in the text where more explanation would be helpful for non-experts (see recommendations).

Reviewer #2 (Public review):

Summary:

Wolbachia are maternally transmitted bacteria that can manipulate host reproduction in various ways. Some Wolbachia induce male killing (MK), where the sons of infected mothers are killed during development. Several MK-associated genes have been identified in Homona magnanima, including Hm-oscar and wmk-1-4, but the mechanistic links between these Wolbachia genes and MK in the native host are still unclear.

In this manuscript, Arai et al. show that Hm-oscar is the gene responsible for Wolbachia-induced MK in Homona magnanima. They provide evidence that Hm-Oscar functions through interactions with the sex determination system. They also found that Hm-Oscar disrupts sex determination in male embryos by inducing female-type dsx splicing and impairing dosage compensation. Additionally, Hm-Oscar suppresses the function of Masc. The manuscript is well-written and presents intriguing findings. The results support their conclusions regarding the diversity and commonality of MK mechanisms, contributing to our understanding of the mechanisms and evolutionary aspects of Wolbachia-induced MK.

Strengths/weaknesses:

(1) The authors found that transient overexpression of Hm-oscar, but not wmk-1-4, in Wolbachia-free H. magnanima embryos induces female-biased sex ratios. These results are striking and mirror the phenotype of the wHm-t infected line (WT12). However, Table 1 lists the "male ratio," while the text presents the "female ratio" with standard deviation. For consistency, the calculation term should be uniform, and the "ratio" should be listed for each replicate.

(2) The error bars in Figure 3 are quite large, and the figure lacks statistical significance labels. The authors should perform statistical analysis to demonstrate that Hm-oscar-overexpressed male embryos have higher levels of Z-linked gene expression.

(3) The authors demonstrated that Hm-Oscar suppresses the masculinizing functions of lepidopteran Masc in BmN-4 cells derived from the female ovaries of Bombyx mori. They should clarify why this cell line was chosen and its biological relevance. Additionally, they should explain the rationale for evaluating the expression levels of the male-specific BmIMP variant and whether it is equivalent to dsx.

(4) Although the authors show that Hm-oscar is involved in Wolbachia-induced MK in Homona magnanima and interacts with the sex determination system in lepidopteran insects, the precise molecular mechanism of Hm-oscar-induced MK remains unclear. Further studies are needed to elucidate how Hm-oscar regulates Homona magnanima genes to induce MK, though this may be beyond the scope of the current manuscript.

Reviewer #3 (Public review):

Summary:

Overall, this is a clearly written manuscript with nice hypothesis testing in a non-model organism that addresses the mechanism of Wolbachia-mediated male killing. The authors aim to determine how five previously identified male-killing genes (encoded in the prophage region of the wHm Wolbachia strain) impact the native host, Homona magnanima moths. This work builds on the authors' previous studies in which:

(1) They tested the impact of these same wHm genes via heterologous expression in Drosophila melanogaster.

(2) They examined the activity of other male-killing genes (e.g., from the wFur Wolbachia strain in its native host: Ostrinia furnacalis moths).

Advances here include identifying which wHm gene most strongly recapitulates the male-killing phenotype in the native host (rather than in Drosophila), and the finding that the Hm-Oscar protein has the potential for male-killing in a diverse set of lepidopterans, as inferred by the cell-culture assays.

Strengths:

Strengths of the manuscript include the reverse genetics approaches to dissect the impact of specific male-killing loci, and the use of a "masculinization" assay in Lepidopteran cell lines to determine the impact of interactions between specific masc and oscar homologs.

Weaknesses:

My major comments are related to the lack of statistics for several experiments (and the data normalization process), and opportunities to make the manuscript more broadly accessible.

The manuscript I think would be much improved by providing more details regarding some of the genes and cross-lineage comparisons. I know some of this is reported in previous publications, but some summary and/or additional analysis would make this current manuscript much more approachable for a broader audience, and help guide readers to specific important findings. For example, a graphic and/or more detail on how the wmk/oscar homologs (within and across Wolbachia strains) differ (e.g., domains, percent divergence, etc) would be helpful for contextualizing some of the results. I recognize the authors discuss this in parts (e.g., lines 223-227), but it does require some bouncing between sections to follow. Similarly, the experiments presented in Figure 4 indicate that Hm-oscar has broad spectrum activity: how similar are the masc proteins from these various lepidopterans? Are they highly conserved? Rapidly evolving? Do the patterns of masc protein evolution provide any hints at how Oscar might be interacting with masc?

It is clear from Figure 1 that the combinations of wmk homologs do not cause male killing on their own. Did the authors test if any of the wmk homologs impact the MK phenotype of oscar? It looks like a previous study tested this in wFur (noted in lines 250-252), but given that the authors also highlight the differences between the wFur-oscar and Hm-oscar proteins, this may be worth testing in this system. Related to this, what is the explanation for why there would be 4 copies of wmk in Hm?

Why are some of the broods male-biased (2/3) rather than ~50:50? (Lines 170-175, Figure 2a). For example, there is a strong male bias in un-hatched oscar-injected and naturally infected embryos, whereas the control uninfected embryos have normal 50:50 sex ratios. It is difficult to interpret the rate of male-killing given that the sex ratios of different sets of zygotes are quite variable.

Figure 2b - it appears there are both male and female bands in the HmOsc male lane. I think this makes sense (likely a partial phenotype due to the nature of the overexpression approach), but this is worth highlighting, especially in the context of trying to understand how much of the MK phenotype might be recapitulated through these methods. Related, there is no negative control for this PCR.

It appears the RNA-seq analysis (Figure 3) is based on a single biological replicate for each condition. And, there are no statistical comparisons that support the conclusions of a shift in dosage compensation. Finally, it is unclear what exactly is new data here: the authors note "The expression data of the wHm-t-infected and non-infected groups were also calculated based on the transcriptome data included in Arai et al. (2023a)" - So, are the data in Figure 3c and 3d a re-print of previous data? The level of dosage compensation inferred by visually comparing the control conditions in 3b and 3d does not appear consistent. With only one biological replicate library per condition, what looks like a re-print of previous data, and no statistical comparisons, this is a weakly supported conclusion.

In Figure 4: There are no statistics to support the conclusions presented here. Additionally, the data have gone through a normalization process, but it is difficult to follow exactly how this was done. The control conditions appear to always be normalized to 100 ("The expression levels of BmImpM in the Masc and Hm-Oscar/Oscar co-transfected cells were normalized by setting each Masc-transfected cell as 100"). I see two problems with this approach:

(1) This has eliminated all of the natural variation in BmImpM expression, which is likely not always identical across cells/replicates.

(2) How then was the percentage of BmImpM calculated for each of the experimental conditions? Was each replicate sample arbitrarily paired with a control sample? This can lead to very different outcomes depending on which samples are paired with each other. The most appropriate way to calculate the change between experimental and control would be to take the difference between every single sample (6 total, 3 control, 3 experimental) and the mean of the control group. The mean of the control can then be set at 100 as the authors like, but this also maintains the variability in the dataset and then eliminates the issue of arbitrary pairings. This approach would also then facilitate statistical comparisons which is currently missing.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

Insects and their relatives are commonly infected with microbes that are transmitted from mothers to their offspring. A number of these microbes have independently evolved the ability to kill the sons of infected females very early in their development; this male killing strategy has evolved because males are transmission dead-ends for the microbe. A major question in the field has been to identify the genes that cause male killing and to understand how they work. This has been especially challenging because most male-killing microbes cannot be genetically manipulated. This study focuses on a male-killing bacterium called Wolbachia. Different Wolbachia strains kill male embryos in beetles, flies, moths, and other arthropods. This is remarkable because how sex is determined differs widely in these hosts. Two Wolbachia genes have been previously implicated in male-killing by Wolbachia: oscar (in moth male-killing) and wmk (in fly male-killing). The genomes of some male-killing Wolbachia contain both of these genes, so it is a challenge to disentangle the two.

This paper provides strong evidence that oscar is responsible for male-killing in moths. Here, the authors study a strain of Wolbachia that kills males in a pest of tea, Homona magnanima. Overexpressing oscar, but not wmk, kills male moth embryos. This is because oscar interferes with masculinizer, the master gene that controls sex determination in moths and butterflies. Interfering with the masculinizer gene in this way leads the (male) embryo down a path of female development, which causes problems in regulating the expression of genes that are found on the sex chromosomes.

We would like to thank you for evaluating our manuscript.

Strengths:

The authors use a broad number of approaches to implicate oscar, and to dissect its mechanism of male lethality. These approaches include:

(1) Overexpressing oscar (and wmk) by injecting RNA into moth eggs.

(2) Determining the sex of embryos by staining female sex chromosomes.

(3) Determining the consequences of oscar expression by assaying sex-specific splice variants of doublesex, a key sex determination gene, and by quantifying gene expression and dosage of sex chromosomes, using RNASeq.

(4) Expressing oscar along with masculinizer from various moth and butterfly species, in a silkmoth cell line.

This extends recently published studies implicating oscar in male-killing by Wolbachia in Ostrinia corn borer moths, although the Homona and Ostrinia oscar proteins are quite divergent. Combined with other studies, there is now broad support for oscar as the male-killing gene in moths and butterflies (i.e. order Lepidoptera). So an outstanding question is to understand the role of wmk. Is it the master male-killing gene in insects other than Lepidoptera and if so, how does it operate?

Thank you for your comments. Wolbachia strains often carry wmk genes, but as observed in this study, the homologs in Homona showed no apparent MK ability. These showed strong male lethality in D. melanogaster, but it is still unclear whether the genes are the master male-killing gene in Diptera. It is also possible that the genes show toxicities in other lepidopteran insects as well as in other insect taxa. Further functional validation assays in different insects are warranted to clarify whether wmk shows toxicity in different insect taxa. We have also discussed the functions of wmk in the Discussion section (lines 301-306).

Weaknesses:

I found the transfection assays of oscar and masculinizer in the silkworm cell line (Figure 4) to be difficult to follow. There are also places in the text where more explanation would be helpful for non-experts (see recommendations).

Thank you for your suggestion. We have thoroughly revised the manuscript to address all the questions, comments and suggestions you raised in “recommendations”. In particular, we have revised the section on the transfection assays of Oscar and Masc in Bm-N4 cells (result section “Hm-oscar suppresses the masculinizing functions of lepidopteran masc genes” starts on line 214 and Fig. 4; materials and methods section ”Transfection assays and quantification of BmIMPM”, starts on line 483). We have also provided more detailed explanations for non-experts in some contexts (in response to your recommendation). We believe that the resulting revisions have significantly improved the quality and comprehensiveness of our manuscript.

Reviewer #2 (Public review):

Summary:

Wolbachia are maternally transmitted bacteria that can manipulate host reproduction in various ways. Some Wolbachia induce male killing (MK), where the sons of infected mothers are killed during development. Several MK-associated genes have been identified in Homona magnanima, including Hm-oscar and wmk-1-4, but the mechanistic links between these Wolbachia genes and MK in the native host are still unclear.

In this manuscript, Arai et al. show that Hm-oscar is the gene responsible for Wolbachia-induced MK in Homona magnanima. They provide evidence that Hm-Oscar functions through interactions with the sex determination system. They also found that Hm-Oscar disrupts sex determination in male embryos by inducing female-type dsx splicing and impairing dosage compensation. Additionally, Hm-Oscar suppresses the function of Masc. The manuscript is well-written and presents intriguing findings. The results support their conclusions regarding the diversity and commonality of MK mechanisms, contributing to our understanding of the mechanisms and evolutionary aspects of Wolbachia-induced MK.

We would like to thank you for evaluating our manuscript.

Strengths/weaknesses:

(1) The authors found that transient overexpression of Hm-oscar, but not wmk-1-4, in Wolbachia-free H. magnanima embryos induces female-biased sex ratios. These results are striking and mirror the phenotype of the wHm-t infected line (WT12). However, Table 1 lists the "male ratio," while the text presents the "female ratio" with standard deviation. For consistency, the calculation term should be uniform, and the "ratio" should be listed for each replicate.

We have revised the first results section (Hm-oscar induces female-biased sex ratios, starting from line 147) accordingly to maintain the consistency in the calculation term. In the revised manuscript, the 'male ratio' is now consistently used, in alignment with Fig. 1. In addition, we have included all sex ratio information (number of males and females) in the supplementary data file for transparency and clarity.

(2) The error bars in Figure 3 are quite large, and the figure lacks statistical significance labels. The authors should perform statistical analysis to demonstrate that Hm-oscar-overexpressed male embryos have higher levels of Z-linked gene expression.

The large error bar on each chromosome (Fig.3a-d) likely reflect the overall variation in expression levels across different transcripts. Accordingly, we have included statistical data for Figure 3 based on the Steel-Dwass test for expression levels. However, displaying statistical significance directly on the whisker plots would make the figure too cluttered due to the numerous combinations. Instead, we have provided all the statistical data in the supplementary data file. To further support the claim that Z-linked genes are more highly expressed in wHm-t-infected/Hb-Oscar-injected embryos, we have included the expression data for a Z-linked gene tpi, along with its statistical data in the revised manuscript (Fig. 3e, lines 210-212).

(3) The authors demonstrated that Hm-Oscar suppresses the masculinizing functions of lepidopteran Masc in BmN-4 cells derived from the female ovaries of Bombyx mori. They should clarify why this cell line was chosen and its biological relevance. Additionally, they should explain the rationale for evaluating the expression levels of the male-specific BmIMP variant and whether it is equivalent to dsx.

Thank you for your suggestion. We selected BmN-4 cell line because previous studies have established it as a reliable model for investigating the functions of lepidopteran masc genes and the interactions between masc and Oscar genes (Katsuma et al., 2019; 2022). In addition, BmIMPM is a male-specific regulator of the male-type dsx, making it an ideal target for assessing the 'maleness' induced by transfection of the masc gene in female-derived BmN-4 cells (Suzuki et al., 2010; Katsuma et al., 2015). We have included more detailed background information in the revised manuscript and have thoroughly revised this section (Hm-oscar suppresses the masculinizing functions of lepidopteran masc genes, starting at line 214) and Figure 4 for better clarity.

(4) Although the authors show that Hm-oscar is involved in Wolbachia-induced MK in Homona magnanima and interacts with the sex determination system in lepidopteran insects, the precise molecular mechanism of Hm-oscar-induced MK remains unclear. Further studies are needed to elucidate how Hm-oscar regulates Homona magnanima genes to induce MK, though this may be beyond the scope of the current manuscript.

Based on our findings and previous studies in Homona, Ostrinia and Bombyx (Arai et al., 2023a; Katsuma et al., 2023; Kiuchi et al., 2014), we hypothesize that the molecular mechanisms underlying _w_Hm-induced MK are likely linked to impaired dosage compensation caused by the inhibition of Masc function by the Hm-Oscar protein. While the precise mechanisms remain unclear, unbalanced Z-linked gene expression due to the impaired dosage compensation (i.e., 2-fold higher Z-linked gene expression compared to normal males) is known to be lethal for lepidopteran males (Kiuchi et al., 2014; Fukui et al., 2015; Visser et al., 2021). We have outlined this hypothesis in the Discussion section (lines 245-254).

Reviewer #3 (Public review):

Summary:

Overall, this is a clearly written manuscript with nice hypothesis testing in a non-model organism that addresses the mechanism of Wolbachia-mediated male killing. The authors aim to determine how five previously identified male-killing genes (encoded in the prophage region of the wHm Wolbachia strain) impact the native host, Homona magnanima moths. This work builds on the authors' previous studies in which:

(1) They tested the impact of these same wHm genes via heterologous expression in Drosophila melanogaster.

(2) They examined the activity of other male-killing genes (e.g., from the wFur Wolbachia strain in its native host: Ostrinia furnacalis moths).

Advances here include identifying which wHm gene most strongly recapitulates the male-killing phenotype in the native host (rather than in Drosophila), and the finding that the Hm-Oscar protein has the potential for male-killing in a diverse set of lepidopterans, as inferred by the cell-culture assays.

Strengths:

Strengths of the manuscript include the reverse genetics approaches to dissect the impact of specific male-killing loci, and the use of a "masculinization" assay in Lepidopteran cell lines to determine the impact of interactions between specific masc and oscar homologs.

We would like to thank you for evaluating our manuscript.

Weaknesses:

My major comments are related to the lack of statistics for several experiments (and the data normalization process), and opportunities to make the manuscript more broadly accessible.

Thank you for your suggestions. We have thoroughly revised the manuscript to provide clearer explanations for non-experts. In addition, we have included more detailed statistical data for Figure 3 and Figure 4 based on the Steel-Dwass tests. For Figure 3a-d, displaying statistical significance directly on the whisker plots would make the figure too cluttered due to the numerous combinations. Therefore, we have provided all the statistical data in the supplementary data file. To further support the claim that Z-linked genes are more highly expressed in _w_Hm-t-infected/Hm-Oscar-injected embryos, we have included the expression data for a Z-linked gene tpi, along with its statistical data in the revised manuscript (Fig.3e, lines 210-212). Regarding Figure 4, we have revised the Figure based on the reviewer’s suggestions, and provided more detailed information on how the expression data were analyzed (Transfection assays and quantification of BmIMPM, lines 495-520). We have also included more detailed background information on the assay system (Hm-oscar suppresses the masculinizing functions of lepidopteran masc genes, lines 215-237). Although we did not observe statistical significance based on the Steel-Dwass test, likely due to limited replicates, the observed changes in the IMP gene expression remain clearly evident.

The manuscript I think would be much improved by providing more details regarding some of the genes and cross-lineage comparisons. I know some of this is reported in previous publications, but some summary and/or additional analysis would make this current manuscript much more approachable for a broader audience, and help guide readers to specific important findings. For example, a graphic and/or more detail on how the wmk/oscar homologs (within and across Wolbachia strains) differ (e.g., domains, percent divergence, etc) would be helpful for contextualizing some of the results. I recognize the authors discuss this in parts (e.g., lines 223-227), but it does require some bouncing between sections to follow. Similarly, the experiments presented in Figure 4 indicate that Hm-oscar has broad spectrum activity: how similar are the masc proteins from these various lepidopterans? Are they highly conserved? Rapidly evolving? Do the patterns of masc protein evolution provide any hints at how Oscar might be interacting with masc?

Thank you for your valuable suggestion. To address this, we have included a visualization of the structural differences between the Oscar and wmk homologs in Figure 1a of the revised manuscript. In addition, we have included more detailed information for these genes and revised the introduction (lines 110-114; 124-137) and discussion (lines 255-266) to provide a clearer and more comprehensive overview. We have also described the similarity of the Masc proteins and Oscar proteins that we used, which is now reflected in the revised Figure 4b and 4d. More detailed information on these proteins is available in the supplementary data. Notably, Masc proteins exhibit high sequence variability with conserved domains (Figure 4d). Previous study identified the N-terminal region of Masc as crucial for the Oscar function (Katsuma et al., 2022). The wide spectrum of the actions of Hm-Oscar likely stems from these conserved structures of Masc, but the effects might have undergone evolutionary tuning through interactions with the native host as discussed in lines 293-294.

It is clear from Figure 1 that the combinations of wmk homologs do not cause male killing on their own. Did the authors test if any of the wmk homologs impact the MK phenotype of oscar? It looks like a previous study tested this in wFur (noted in lines 250-252), but given that the authors also highlight the differences between the wFur-oscar and Hm-oscar proteins, this may be worth testing in this system. Related to this, what is the explanation for why there would be 4 copies of wmk in Hm?

Thank you for your valuable suggestion. Unfortunately, we have not yet tested the effects of co-expression of wmk and Oscar. Due to a technical issue, the mixing of multiple constructs results in a reduced amount of mRNA (i.e. mixing wmk-3 and Hm-Oscar at the same concentration results in a 2-fold lower concentration in mRNA for both genes compared to mono-injected groups). In addition, we have previously tested injecting mRNA at the twofold higher concentration (i.e. 2 ug/ul mRNA), which resulted in very low hatchability regardless of the genes. Katsuma et al (2022) tested the effect of wmk on the sex determination system, but did not test the effect of co-injection/transfection of wmk and Oscar. Considering the results of this and previous studies (Katsuma et al., 2022; Arai et al., 2023), it is likely that the targets of the wmk and oscar genes are different (as discussed in lines 267-289). Co-injection of wmk and oscar may not produce additive effects. Nevertheless, we would like to test the results in future studies using the Drosophila system as well.

As you point out, it is an interesting point that the moth-derived MK Wolbachia _w_Hm-t encodes four wmk genes, although they have no apparent effect on host survival. The exact functional relevance of these wmk homologs remains unclear. However, they may play a role in Wolbachia biology as transcriptional regulators, given that they encode HTH domains. Wolbachia generally encode several wmk homologs in their genome, regardless of whether they induce MK. This suggests that the functions of the wmk genes may be 'suppressed' in certain Wolbachia-host systems. The wmk and Hm-oscar genes are located within a prophage region, and some wmk genes are tandemly arrayed (as described in Arai et al., 2023). These wmk homologs may have increased in number by horizontal phage transfer, and the region containing wmk and adjacent sequences may act as a genomic island for virulence. So far, the function of wmk homologs has only been tested in D. melanogaster and H. magnanima, and further studies in other Wolbachia-host systems are highly warranted to test whether wmk exerts MK effects in other insect models. These points have been briefly discussed in the revised manuscript (lines 301-306; 318-320).

Why are some of the broods male-biased (2/3) rather than ~50:50? (Lines 170-175, Figure 2a). For example, there is a strong male bias in un-hatched oscar-injected and naturally infected embryos, whereas the control uninfected embryos have normal 50:50 sex ratios. It is difficult to interpret the rate of male-killing given that the sex ratios of different sets of zygotes are quite variable.

The observed male-biased sex ratios in unhatched embryos are due to the occurrence of MK during embryogenesis. In the unhatched groups, the skew towards males reflects that fact that the male embryos were targeted and killed by Wolbachia/Oscar, leading to a surplus of unhatched male embryos. Conversely, hatched individuals show a higher proportion of females because many of the males were eliminated during embryogenesis. Thus, the unhatched embryos are more male-biased, while the hatched individuals are more female-biased in the Hm-oscar/_w_Hm-t treated groups. We have revised the relevant section (Males are killed mainly at the embryonic stage, lines 179-186) and provided more detailed information to clarify this explanation.

Figure 2b - it appears there are both male and female bands in the HmOsc male lane. I think this makes sense (likely a partial phenotype due to the nature of the overexpression approach), but this is worth highlighting, especially in the context of trying to understand how much of the MK phenotype might be recapitulated through these methods. Related, there is no negative control for this PCR.

Thank you for your suggestion. As you noted, a faint dsx-M band is visible in the Hm-oscar treated group in Figure 2b. This is consistent with previous findings by Arai et al. (2023), which reported that male embryos with low-density _w_Hm-t showed double bands of dsx-M and dsx-F, similar to what we observed in this study. This information has been included in the revised manuscript in lines 196-198, as follows:

“Notably, male embryos expressing Hm-oscar also exhibited weak male-type dsx splicing in addition to the female-type splicing, resembling the previously observed pattern in male embryos infected with low-titer _w_Hm-t (Arai et al., 2023a).”

Also, we appreciate your comment regarding the missing of negative control. The figure has now been revised as we realised that the negative control lane had been lost during the preparation of the figure. We also included the relevant molecular marker information in both the figure legends and Figure 2b.

It appears the RNA-seq analysis (Figure 3) is based on a single biological replicate for each condition. And, there are no statistical comparisons that support the conclusions of a shift in dosage compensation. Finally, it is unclear what exactly is new data here: the authors note "The expression data of the wHm-t-infected and non-infected groups were also calculated based on the transcriptome data included in Arai et al. (2023a)" - So, are the data in Figure 3c and 3d a re-print of previous data? The level of dosage compensation inferred by visually comparing the control conditions in 3b and 3d does not appear consistent. With only one biological replicate library per condition, what looks like a re-print of previous data, and no statistical comparisons, this is a weakly supported conclusion.

Thank you for your suggestion. In this study, we generated the RNA-seq data for the Hm-oscar/GFP-injected groups, but did not sequence the _w_Hm-t-infected/NSR lines. Instead, the previously generated RNA-seq data of _w_Hm-t-infected/NSR (Arai et al., 2023) were re-analyzed (rather than simply reprinted) to evaluate whether the expression patterns of Hm-oscar-injected and _w_Hm-t-infected groups are similar. We have revised the Results section (Hm-oscar impairs dosage compensation in male embryos, lines 200-212), the Materials and methods section (Quantification of Z chromosome-linked genes, lines 454-456), and the figure legends to provide more precise information about this analysis.

Although we did not perform replicates for the RNA-seq comparisons, it is important to note that each RNA-seq sample contains 50-60 male/female individuals. We believe the results are still robust and clearly indicative of the trends we observe. This was further supported by the quantification of Hmtpi gene expression, which we have visualized in Figure 3e (and lines 210-212). As you noted, the expression patterns in Figure 3b (GFP injected) and Figure 3d (NSR) are not completely identical. This discrepancy may be due to the differences between injection treatments and natural infections. Nevertheless, both treatments are consistent in showing that gene expressions on the Z chromosome (Chr01 and Chr15) are not upregulated.

We have also added more detailed statistical data for Figure 3 based on the Steel-Dwass tests. For Figure 3a-d, however, showing the statistical significance directly on the whisker plots would create excessive clutter due to the numerous combinations of chromosomes. Instead, we have provided the full statistical data in the supplementary data file. Furthermore, to support/strengthen our conclusion that Z-linked genes are highly expressed in _w_Hm-t-infected/Hm-Oscar-injected embryos, we have included expression data for the Z-linked gene tpi, along with statistical data, in the revised manuscript (Fig. 3e, lines 210-212).

In Figure 4: There are no statistics to support the conclusions presented here. Additionally, the data have gone through a normalization process, but it is difficult to follow exactly how this was done. The control conditions appear to always be normalized to 100 ("The expression levels of BmImpM in the Masc and Hm-Oscar/Oscar co-transfected cells were normalized by setting each Masc-transfected cell as 100"). I see two problems with this approach:

(1) This has eliminated all of the natural variation in BmImpM expression, which is likely not always identical across cells/replicates.

(2) How then was the percentage of BmImpM calculated for each of the experimental conditions? Was each replicate sample arbitrarily paired with a control sample? This can lead to very different outcomes depending on which samples are paired with each other. The most appropriate way to calculate the change between experimental and control would be to take the difference between every single sample (6 total, 3 control, 3 experimental) and the mean of the control group. The mean of the control can then be set at 100 as the authors like, but this also maintains the variability in the dataset and then eliminates the issue of arbitrary pairings. This approach would also then facilitate statistical comparisons which is currently missing.

Thank you for your suggestion. As you pointed out in (1), the previous analysis did indeed eliminate the natural variation in BmIMP-M expression. In the revised manuscript and Figure 4, we have reanalyzed the data following your suggestion and have described the variation across replicates.

For (2), the data shown in the previous manuscript were normalized to 100 for each Masc-treated group. In doing so, each replicate sample was arbitrarily paired with a control sample from the same cell lot to account for variations that might occur due to differences in cell lots. However, following your recommendation, we have revised the figure to set the average of the Hm-masc treated group to 100, rather than using arbitrary pairings. More detailed normalization procedures have been provided in the section 'Transfection assays and quantification of BmIMP' (lines 483-520). Additionally, we have provided more detailed background information on the assay system in lines 218-223. Although we did not observe statistical significance based on the Steel-Dwass test, likely due to the limited number of replicates, the differences in IMP gene expression between the Masc-treated and Masc&Hm-oscar-treated groups remain evident.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Line 38: change to: 'Wolbachia are maternally transmitted'.

Revised accordingly (line 38).

Line 69: remove 'seemingly'.

Revised accordingly (line 69).

Paragraph starting line 123: I don't think this is so clear to a reader who is not familiar with the work and system. It would be helpful to more clearly explain that candidate male-killing genes from Wolbachia that infect Homona were inserted into Drosophila melanogaster, and that their expression was then induced, with interesting patterns (and that it can be a bit difficult to interpret the transgenic expression of genes from a moth male-killer that are inserted into a fly). Also, the sentence about the combined action of cifA and cifB in Drosophila cytoplasmic incompatibility is also confusing to a non-expert. I would suggest removing it.

Thank you for your suggestion. We have revised the paragraph (lines 124-139) to provide clearer background information, making it easier for non-experts to follow. We have also removed the sentence regarding the combined effect of cifA and cifB to improve the flow and overall clarity.

Line 170: what is the explanation for the male-biased sex ratio instead of 50-50?

The male-biased sex ratio occurs because MK happens during embryogenesis. Unhatched embryos include males that were killed by Wolbachia/Oscar, resulting in a higher proportion of unhatched male embryos. Conversely, the hatched individuals display a female bias, as most of the males were eliminated during embryogenesis. Thus, the unhatched embryos are more male-biased, while the hatched individuals are more female-biased in the Hm-oscar/_w_Hm-t treated groups. We have revised the section “Males are killed mainly at the embryonic stage” (lines 170-186) to include more detailed information explaining this phenomenon.

Line 190: please explain what are the Z chromosomes in Bombyx and Homona and Lepidoptera in general (chromosomes 1 and 15?), as this is not so clear for a non-expert.

Thank you for your suggestion. I have revised the section (lines 200-212) to include more precise background information about the chromosome constitutions in lines 202-204 as follows:

“Unlike other lepidopteran species, Tortricidae, including H. magnanima, generally possess a large Z chromosome that is homologous to B. mori chromosomes 1 (Z) and 15 (autosome).”

Line 222: please explain oscar diversity and classification in more detail, as this is not so clear for a non-expert.

Thank you for your suggestion. We have revised the sentences to provide clearer background information on the diversity of oscar genes (lines 255-264).

Figure 4: I found this difficult to follow. Why are there 2 rows (HmOscar and Oscar)? Does oscar here refer to oscar from Ostrinia? I am also a bit confused about the baseline control of Masc in these cell lines. If I understand Lepidoptera sex determination, then these cell lines are expressing high levels of female-specific piRNAs that suppress Masc. How specific are these piRNAs (i.e. do Bombyx piRNAs suppress Mascs from other Lepidoptera)? How much extra Masc will override endogenous piRNA? Information is lost by setting Masc expression to 100% in each separate comparison.

Yes, the Oscar indicates the _w_Fur-encoded oscar (Oscar from Ostrinia) that was tested to compare function with the Homona-derived Hm-oscar gene. In addition, following the reviewer's suggestions, we have revised the figure and included more detailed information on how we adjusted the expressions in the M&M section.

A previous study (Shoji et al., 2017, RNA 23:86–97) demonstrated that the Fem piRNA (29 bp) in Bombyx mori requires a 17 bp complementary sequence from its 5' region for its function. However, in species other than B. mori, no significant homology (i.e., over 17 bp matches) was found between the B. mori Fem piRNA and the masc genes analyzed in this study. Therefore, it is likely that the Fem piRNA expressed in BmN-4 cells is unable to suppress the masculinizing function driven by masc genes in other lepidopteran species. In addition, we did not quantify the levels of piRNA in this system, but the expression levels of masc are probably too high to be suppressed.

Figure 4 legend: spelling of Spodoptera.

Revised accordingly.

Reviewer #2 (Recommendations for the authors):

In Figure 2, what is the dsx splicing type for the hatched male in the Hm-oscar-injected group and the wHm-t infected line? Dsx-F or dsx-M?

Thank you for your suggestion. Unfortunately, we have not tested splicing in the hatched male neonates (1st instar larvae), partly due to difficulties in obtaining sufficient material for RNA extraction. Based on the previous publication in the Ostrinia system, where Oscar-bearing _w_Sca induces MK, the hatched males (ZZ) exhibit female type dsx as observed in the male embryos (Herran et al., 2022). The hatched Homona males may show double bands for dsx-M and dsx-F as observed in this study.

The size of the markers (in kilobase pairs) should be indicated in Figure 2.

We have accordingly included the marker information in the revised Figure 2b and the figure legends.

In Figure 3, could the authors identify which genes exhibit higher expression levels in the Hm-oscar-injected group and the wHm-t infected line? Could they provide hints for the possible mechanism of male-killing?

In the RNA-seq data shown in Figure 3a-d, we observed that both the Hm-oscar-injected and _w_Hm-infected groups generally exhibited upregulated expression of Z-linked genes. Rather than the upregulation or downregulation of a specific gene, we consider that global upregulation of Z-linked genes, caused by improper dosage compensation, is lethal for males. The Z chromosome contains various genes involved in key biological processes such as endocrine function and detoxification, and disruption of these processes may contribute to male lethality. Additionally, in this revised manuscript, we have provided more detailed information on the expression level of the Z-linked gene tpi. We have also discussed the potential mechanisms of MK in the Discussion section (lines 245-254).

The format of the references should be consistent. Gene and species names should be italicized.

We have accordingly formatted.

Reviewer #3 (Recommendations for the authors):

The authors use the term "upstream" (e.g., Oscar suppressed the function of masculinizer, the upstream male sex determinant...), which was sometimes confusing. In many cases, it reads as though the masculinizer was upstream of oscar, but what I think the authors are trying to convey is that masculinizer is a primary sex-determining factor.

Thank you for your suggestion. We have accordingly revised the term.

Line 101: which insect is wFur from?

It is from Ostrinia furnacalis - line 104 has been revised.

Figure 1: it would be helpful to indicate the statistical results on the figure.

Accordingly, we have added statistical data (binominal test) for Figure 1. The data for the Steel-Dwass test have been included in the supplementary data.

Figure 2b: please label the ladder on the gel.

Thank you for your suggestion. We have accordingly labeled the DNA ladder on the gel.

Version 2

Abstract

Abstract

Wolbachia are maternally transmitted bacterial symbionts that are ubiquitous among arthropods. They can hijack host reproduction in various ways, including male killing (MK), where the sons of infected mothers are killed during development. The recent discovery of MK-associated Wolbachia genes, i.e., oscar in Ostrinia moths and wmk in Drosophila flies, stimulates our interest in the diversity and commonality of MK mechanisms, which remain largely unclear. We recently discovered that a Wolbachia symbiont of the moth Homona magnanima carries an MK-associated prophage region encoding homologs of oscar (Hm-oscar) and wmk (wmk-1–4). Here, we investigated the effects of these genes in the native host. Upon transient overexpression, Hm-oscar, but not wmk, induced male lethality in H. magnanima, in contrast to our observations in Drosophila, where the wmk homologs, but not Hm-oscar, killed the males. Hm-oscar disrupted sex determination in male embryos by inducing a female-type doublesex splicing and impaired dosage compensation, recapitulating the Wolbachia phenotype. Cell-based transfection assays confirmed that Hm-oscar suppressed the function of masculinizer, the primary male sex determinant involved in lepidopteran dosage compensation. Our study highlights the conserved roles of oscar homologs in Wolbachia-induced lepidopteran MK and argues that Wolbachia have evolved multiple MK mechanisms in insects.

Introduction

Arthropods commonly carry microbial symbionts that are passed from mother to offspring (Hurst & Frost, 2015; Werren et al., 2008; Hurst, 2017). The maternally transmitted bacteria, belonging to the genus Wolbachia (Alphaproteobacteria), are estimated to be present in at least 40% of all insect species, making them one of the most widespread endosymbiont genera on the planet (Werren et al, 2008; Zug & Hammerstein, 2012). Wolbachia have achieved evolutionary success by manipulating host reproduction through various means that enhance endosymbiont transmission (Werren et al., 2008). Such manipulation of the host’s reproduction includes cytoplasmic incompatibility (CI), parthenogenesis, feminization, and male killing (MK), each of which affects the biological features, distribution, and evolution of the host. Among these strategies, MK directly skews the sex ratio of the host population toward females by killing male offspring of infected mothers during development. The lack of symbiont transmission through male hosts often leads to the evolution of MK. Wolbachia have been shown to induce MK in a wide range of insect taxa (e.g., dipterans, lepidopterans, and coleopterans) and other arthropods (Hurst et al., 1999; Kageyama & Traut, 2004; Hurst et al., 2000). Furthermore, various bacteria, viruses, and microsporidia induce MK phenotypes (Kageyama et al., 2012; Fujita et al., 2021; Kageyama et al., 2023; Nagamine et al., 2023), and recent studies have postulated that these microbes have evolved their MK ability independently (Harumoto & Lemaitre, 2018; Kageyama et al., 2023; Nagamine et al., 2023; Arai et al., 2023a).

The evolution and molecular mechanisms underlying Wolbachia-induced MK have attracted considerable attention for decades. It has been hypothesized that the mechanisms underlying MK are associated with sex determination cascades in insects (Hornett et al., 2022). Indeed, some MK-inducing Wolbachia strains have been shown to directly interact with the sex determination system in lepidopteran insects (Sugimoto and Ishikawa, 2012; Sugimoto et al., 2015; Fukui et al., 2015; Arai et al., 2023a). The Lepidopteran sex determination system consists of multiple transcriptional regulators, some of which exhibit sex-linked expression and/or splicing isoforms. Lepidoptera generally have a female heterogametic sex chromosome system (e.g., WZ/ZZ) and employ dosage compensation, which equalizes the sex-linked (Z-chromosome-linked) gene dose between males and females (Kiuchi et al., 2014). Dosage compensation is regulated by the masculinizer gene (masc), which is critical for male development. masc also controls the downstream master regulator of sex determination and differentiation, doublesex (dsx), which exhibits sex-dependent splicing isoforms (dsxF in females and dsxM in males) (Kiuchi et al., 2014). In Ostrinia and Homona male moths, MK-inducing Wolbachia strains impair dosage compensation by disrupting the expression of masc. Furthermore, they disrupt sex determination in male moths by inducing the “female” isoform of dsx (dsxF), leading to a “mismatch” between genetic sex (male: ZZ sex chromosome constitution) and phenotypic sex (female: based on dsxF) and consequently to male death (Arai et al., 2023a; Fukui et al., 2015; Sugimoto et al., 2015; Sugimoto & Ishikawa, 2012).

More recently, a protein produced in Wolbachia wFur strain named Oscar (Osugoroshi protein containing CifB C-terminus-like domain and many Ankyrin Repeats; Osugoroshi translates to male killing in Japanese) was shown to recapitulate the Wolbachia-induced MK phenotype in the native host Ostrinia furnacalis (Katsuma et al., 2022). The Oscar protein interacts with and degrades Masc protein, leading to the failure of dosage compensation and the production of female-type dsx isoforms in Ostrinia male moths (Fukui et al., 2024; Katsuma et al., 2022). Although oscar homologs have been identified in several MK-inducing Wolbachia strains in Lepidoptera, some MK Wolbachia strains in Diptera (Drosophila) and Lepidoptera do not carry this gene (Arai et al., 2023b; Arai et al., 2024a). Furthermore, oscar homologs are evolutionarily dynamic, with highly variable sequences and structures (classified as type I to type III oscar based on sequence homology) (Arai et al., 2024a), making it difficult to assess their functional relevance. In addition to oscar, the helix-turn-helix domain-containing putative transcriptional regulator wmk is widely conserved among Wolbachia strains and induces various toxicities in Drosophila flies (i.e., no effect, weak male lethality [30% of males die], strong male lethality [90%–100% of males die], and death of all males and females) (Perlmutter et al., 2019; Perlmutter et al., 2020; Perlmutter et al., 2021; Arai et al., 2023b). Although the mechanisms underlying the wmk-induced toxicities and their connection to sex determination systems remain unclear, these findings suggest that Wolbachia strains carry multiple factors that cause male lethality. However, the diversity and commonality of these functions in insects remain largely unknown, partly due to the technical challenges in validating gene functions in non-model insects.

We recently discovered a prophage region responsible for a Wolbachia-induced MK by comparing closely related non-MK Wolbachia (wHm-c) and MK Wolbachia (wHm-t) in the tea tortrix moth Homona magnanima (Tortricidae) (Arai et al., 2023b; Arai et al., 2024b). The MK-associated prophage element encodes potential MK causative genes–four wmk homologs (wmk-1 to wmk-4) as well as an oscar homolog (Hm-oscar, encoding 1181 aa protein), which differs significantly in sequence length and structure from the wFur-encoded oscar (encoding 1830 aa protein) (Fig. 1a). When these MK candidate genes are transgenically overexpressed in Drosophila, wmk-1 and wmk-3 have a lethality of almost 100%, while wmk-2, wmk-4, and Hm-oscar induce no lethal effects when singly overexpressed. Furthermore, co-expression of the adjacent wmk-3 and wmk-4 has been shown to induce the death of 90% of male flies and restores female survival, suggesting their combined action (Arai et al., 2023b). However, transgenic expression of genes from a moth-derived male killer in flies does not necessarily replicate the native host–bacteria interaction, and the mechanistic links between prophage-encoded Wolbachia genes and MK in the native host Homona remain unclear.

Transient expressions of the Hm-oscar gene resulted in female-biased sex ratios

(a) Structures of wFur-encoded Oscar and wHm-t-encoded Hm-Oscar and four Wmk proteins. (b) Male ratio of adult progeny obtained from cRNA-injected groups (wmk-1, wmk-2, T2A-brigged wmk-1 and wmk-2, wmk-3, wmk-4, T2A-brigged wmk-3 and wmk-4, GFP, and Hm-oscar; n=5–15 independent replicates using different egg masses), the non-injected NSR line, and the MK wHm-t-positive WT12 line. The total numbers of adult females and males are shown at the bottom. Different letters indicate significant differences (Steel–Dwass test, P < 0.05). The dot plots show all data points individually.

In this study, we showed that the prophage-encoded Hm-oscar recapitulates Wolbachia-induced MK in H. magnanima. Furthermore, we clarified the mechanistic links to host sex-determination cascades both in vivo and in vitro and discussed the underlying mechanisms of MK in Lepidoptera, arguing for the diverse evolutionary origin of Wolbachia-induced MK.

Results

Hm-oscar induces female-biased sex ratios

To achieve the transient overexpression of Hm-oscar and the four wmk genes (wmk-1, wmk-2, wmk-3, and wmk-4), constructed mRNA (cRNA) was injected into Wolbachia-free H. magnanima embryos. Subsequently, the adult moths that emerged from the cRNA-injected embryos were sexed based on their external morphology. When Hm-oscar was overexpressed, the sex ratio of adults was strongly female-biased (42 males and 218 females in total, 14.4% ± 11.1% male ratio in 15 replicates), which was in sharp contrast with the ratios observed in the GFP-injected (126 females and 123 males in total, 49.1% ± 2.08% male ratio in 10 replicates) and non-injected (NSR) (201 females and 215 males in total, 51.4% ± 3.08% male ratio in 9 replicates) groups (P = 0.001 and 0.002, respectively, Steel–Dwass test, Fig. 1b). Compared with the GFP-injected group, the sex ratio was not biased by the overexpression of wmk-1 (51.1% ± 7.27% male ratio in 6 replicates, P = 0.999), wmk-2 (53.2% ± 4.77% male ratio in 6 replicates, P = 0.810), wmk-3 (53.3% ± 7.91% male ratio in 5 replicates, P = 0.999), and wmk-4 (52.9% ± 3.47% male ratio in 5 replicates, P = 0.773). Although the dual expression of wmk-3 and wmk-4 induces strong male lethality in Drosophila (Arai et al., 2023b), here, it did not skew the sex ratio of Homona compared with the ratio detected in the GFP-injected group (57.9% ± 7.67% male ratio in 5 replicates, P = 0.704). Similarly, the dual expression of the tandemly arrayed wmk-1 and wmk-2 did not bias the sex ratio (49.4% ± 2.82% male ratio in 5 replicates) compared with the value in the GFP-injected group (P = 1.000). The sex ratio data of all replicates are available on the supplementary data file.

Males are killed mainly at the embryonic stage

The sex of unhatched embryos and hatched larvae (neonates) was determined by karyotyping for W chromatin, with the presence and absence of this substance indicating females and males, respectively. MK Wolbachia wHm-t kills Homona males during embryogenesis, resulting in a female-biased sex ratio in hatched larvae (neonates) and a male-biased sex ratio in unhatched embryos (Arai et al., 2023b; Arai et al., 2020). In line with this finding, the sex ratios of hatched larvae (34 females and 6 males) and unhatched embryos (27 females and 47 males) of the MK wHm-t infected line (WT12) were also female- and male-biased, respectively (P = 0.013 and 0, respectively). In the Hm-oscar-injected group, the sex ratio of the hatched larvae (neonates) was strongly female-biased (21 females and 3 males, P = 0.0001 in the binomial test) (Fig. 2a). In contrast, the unhatched mature embryos showed a male-biased ratio (18 females and 38 males, P = 0.005), suggesting that male mortality primarily occurred during the embryonic stage, reflecting the wHm-t-induced MK phenotype. In the non-injected group (NSR), the sex ratios of hatched larvae (28 females and 28 males) and unhatched embryos (23 females and 25 males) were not biased (P = 0.553 and 0.443, respectively).

Hm-Oscar induced lethality in male embryos with the female-type sex determination

(a) Sex ratio of the hatched larvae and unhatched embryos in the Hm-oscar-expressed, wHm-t-infected, and non-infected/expressed groups. Females and males were discriminated based on the presence or absence of W chromatin. The number of individuals is indicated in brackets. (b) Splicing patterns of the downstream sex-determining gene dsx of H. magnanima embryos (5 days post oviposition). Abbreviations: HmOsc, Hm-oscar injected group; GFP, GFP-injected group; WT12, wHm-t-infected line; NSR, non-infected/injected line. M and F indicate W chromatin-negative (ZZ: male genotype) and W chromatin-positive (ZW: female genotype) mature embryos, respectively. dsx-F and dsx-M represent female and male-specific splicing variants, respectively. L: 100 bp DNA ladder (ExcelBand 100 bp DNA Ladder, SMOBIO Technology, Inc., Hsinchu, Taiwan). 0.1, 0.5, 1.0, and 1.5 kb markers are indicated with arrows. N: negative control (water).

Female-type sex determination in male embryos that are destined to die

Splicing patterns of the downstream sex determinant dsx were assessed in each of the H. magnanima embryos that were Hm-oscar-expressed, GFP-expressed, wHm-t-infected, or non-expressed (i.e., non-injected, NSR). In GFP-expressed and NSR embryos, females and males (which were identified by the presence and absence of W chromatin, respectively) exhibited a female- and male-type splicing variant of dsx, respectively. However, both wHm-t-infected and Hm-oscar-expressed embryos induced female-type dsx splicing regardless of the presence or absence of W chromatin (Fig. 2b). Notably, male embryos expressing Hm-oscar also exhibited weak male-type dsx splicing in addition to the female-type splicing, resembling the previously observed pattern in male embryos infected with low-titer wHm-t (Arai et al., 2023a).

Hm-oscar impairs dosage compensation in male embryos

To confirm the effects of Hm-oscar on dosage compensation, we quantified the expression of Z-linked genes using RNA-seq and qPCR. Unlike other lepidopteran species, the family Tortricidae, including H. magnanima, generally possess a large Z chromosome that is homologous to B. mori chromosomes 1 (Z) and 15 (autosome). RNA-seq analysis revealed that, in Hm-oscar-injected embryos, Z-linked genes (homologs on the B. mori chromosomes 1 and 15) were more expressed in males than in females (Fig. 3a), which was not observed in the GFP-injected group (Fig. 3b). Similarly, as previously reported by Arai et al. (2023a), high levels of Z-linked gene expression were also observed in wHm-t-infected males, but not in NSR males (Fig. 3c,d). Statistical data are available on the supplementary data file. The high (i.e., doubled) Z-linked gene expression in both Hm-oscar-expressed and wHm-t-infected males was further confirmed by quantification of the Z-linked Hmtpi gene (Fig. 3e).

Hm-oscar-overexpressed male embryos showed higher levels of Z-linked gene expression

(a-d) Normalized expression levels (TPM) and chromosomal distributions of transcripts in H. magnanima embryos. RNA-seq data of embryos (108 hpo) were used to make the following comparisons: Hm-oscar-injected males versus Hm-oscar injected females (a), GFP-injected males versus GFP-injected females (b), WT12 males versus WT12 females (c), and NSR males versus NSR females (d). The chromosome number for each H. magnanima transcript-derived contig was assigned based on Bombyx mori gene models. The X axis represents the chromosome number of B. mori (shown as chr01 to chr28), and chr01 and chr15 (highlighted in orange) correspond to the Z chromosome of H. magnanima (Arai et al., 2023a). (e) Relative expression of a Z-linked tpi gene in H. magnanima groups. Statistical significance is highlighted with different alphabets.

Hm-oscar suppresses the masculinizing functions of lepidopteran masc genes

To confirm whether Hm-oscar suppresses the functions of the primary male sex determinant masc, their interactions were assessed by transfection using BmN-4 cells, which are derived from B. mori ovaries (Grace, 1967) and exhibit a female-type default sex determination (Kiuchi et al., 2014). Previous studies have shown that transfection of BmN-4 cells with masc cDNA from various lepidopteran insects, such as O. furnacalis, induces the expression of BmdsxM and BmIMPM (Fukui et al., 2015; Katsuma et al., 2019; Katsuma et al., 2022). Since BmIMPM regulates the splicing of BmdsxM and is exclusively expressed in males, we selected BmIMPM as a quantitative marker of masculinization, as described in Katsuma et al. (2022). In contrast with the control groups, which were transfected with the non-inserted pIZ/V5-His vector and showed this default sex determination, the cells transfected with the Hmmasc-inserted pIZ/V5-His vector exhibited the male-type sex determination, as indicated by the increased expression of the male-specific splicing variant of the sex-determining gene BmImpM, although this increase was not statistically significant (Fig. 4a). Furthermore, the masculinizing function of Hmmasc was suppressed by co-transfecting the Hmmasc/Hm-oscar-inserted pIZ/V5-His vectors, as manifested by the low expression levels of BmImpM. wFur-encoded Oscar, which differs from Hm-Oscar in structure and amino acid sequence (Fig. 4b), also suppressed Hmmasc, the function of this gene was less active than that of Hm-oscar (Fig. 4a). Hm-oscar also suppressed the masculinizing function of masc genes derived from various lepidopteran insects (i.e., O. furnacalis, Spodoptera frugiperda, Bombyx mori, Trichoplusia ni, and Papilio machaon), which are highly divergent in their amino acid sequences (Fig. 4c,d). These results suggest that Hm-oscar has a broad spectrum of activity in Lepidoptera.

Hm-oscar suppressed the masculinizing function of lepidopteran Masc

(a) Relative expression levels of the male-specific BmImpM variant in transfected BmN-4 cells. The relative expression of BmImpM in the control (non-inserted pIZ/V5) and Hm-oscar/oscar co-transfected groups was normalized by setting the mean in the Hmmasc single transfected group as 100. (b) Protein structures and homologies of Hm-Oscar and Oscar proteins. (c) Relative expression levels of the BmImpM in Hm-oscar/masc transfected BmN-4 cells. The relative expression of BmImpM in each condition was normalized by setting the mean in the Hmmasc single transfected group as 100. The dot plots show all data points individually. Each experimental condition was replicated three times. Abbreviations: Hm, Homona magnanima; Bm, Bombyx mori; Of, Ostrinia furnacalis; Pm, Papilio machaon; Sf, Spodoptera frugiperda; and Tv, Trilocha varians. (d) Phylogeny and homologies of the Masc protein sequences used in this study.

Discussion

In this study, we showed that the phage-encoded Hm-oscar, but not wmk, induced male lethality and a female-biased sex ratio in H. magnanima. Furthermore, the overexpressed Hm-oscar impaired male sex determination in Homona, recapitulating the wHm-t-induced phenotypes. Cell-based assays confirmed that Hm-oscar suppressed the masculinizing functions of masc. These results strongly suggested that Hm-oscar underlies the wHm-t-induced MK function in H. magnanima. In light of our results and previous findings in Homona, Ostrinia, and Bombyx (Arai et al., 2023a; Katsuma et al., 2022; Kiuchi et al., 2014), we hypothesize the following molecular mechanisms for wHm-t-induced MK: The Hm-Oscar protein targets and inhibits the function of the HmMasc protein, thereby resulting in female sex determination and disrupting dosage compensation, which ultimately leads to male death. Although the full mechanisms remain unclear, unbalanced Z-linked gene expression mediated by improper dosage compensation—resulting in Z-linked gene expression levels twice as high as those in normal males—appear to be lethal for lepidopteran males (Kiuchi et al., 2014; Fukui et al., 2015; Visser et al., 2021).

The Wolbachia-encoded oscar homologs identified so far show diversity in sequence length and are classified as type I, II, and III based on their amino acid sequence homologies. Type I and II Oscar proteins, derived from MK-inducing Wolbachia, contain many ankyrin repeats and a proteinase domain (Arai et al., 2024a). Although long ankyrin repeat sequences are postulated to be critical for the function of wFur-encoded Oscar protein (1830 aa, type I) (Katsuma et al., 2022), our study revealed that the functions of Hm-Oscar protein (1181 aa, type II), which encodes fewer ankyrin repeats, were comparable with those of Oscar protein carried by wFur. Interestingly, type II oscar gene is also present in the feminizing Wolbachia wFem in Eurema butterflies (Arai et al., 2024a). oscar homologs, which inhibit the masculinizing function and induce female sex determination, may have a conserved function in Wolbachia-induced MK and feminization in Lepidoptera.

In contrast to the results of this study, we have previously demonstrated that the phage-encoded wmk, but not Hm-oscar, induces male lethality in D. melanogaster (Arai et al., 2023b). Although the means of expression are different (i.e., transient in Homona and transgenic in Drosophila), this finding highlighted the differences in the mode/range of action of Wolbachia genes between insect species. It has been hypothesized that microbes induce MK in insects by targeting molecular mechanisms involved in sex determination and differentiation (Hornett et al., 2022). Sex determination systems in insects are diverse; for example, Lepidoptera (including H. magnanima) and Diptera (including D. melanogaster) do not share any known sex-determining genes other than dsx (Suzuki, 2018). The different outcomes in Homona and Drosophila are probably due to their different sex determination systems. Because oscar interacts with and suppresses masc, Hm-oscar could induce mortality in Homona males that possess masc, but not in Drosophila males that lack it. Considering that oscar homologs are not present in known MK Wolbachia strains in dipteran insects (Arai et al., 2024c; Katsuma et al., 2022), the mechanisms (i.e., causative genes) of Wolbachia-induced MK probably differ between insect taxa (e.g., between Lepidoptera and Diptera). While the mechanisms underlying wmk-induced lethality remain unclear, the distinct effects associated with this gene between Homona and Drosophila may also reflect their genetic backgrounds (e.g., presence/absence of host factor(s) that interact with wmk). In addition, Katsuma et al. (2022) reported that the wmk homologs encoded by wFur did not affect the masculinizing function of masc in vitro, indicating that wmk likely targets factors other than masc. Our results strongly suggested that Wolbachia strains have acquired multiple MK genes through evolution.

Wolbachia-induced phenotypes are known to be influenced by the genetic backgrounds of hosts (Hornett et al., 2006; Sasaki et al., 2002; Veneti et al., 2012). Our study showed that the wHm-t-encoded Hm-oscar suppresses the function of masc in H. magnanima more efficiently than the wFur-encoded oscar, suggesting that this Wolbachia factor has undergone evolutionary tuning to adapt to its natural host. However, the mere presence of oscar and wmk homologs does not ensure MK expression. For example, wCauA, bearing type I oscar, did not induce MK in Ephestia (Cadra) cautella collected around 2000, although it did induce MK when transferred to the closely related host Ephestia kuehniella (Sasaki et al., 2002). An MK phenotype, presumably induced by wCauA, was also recorded in E. cautella around the 1970s (Takahashi & Kuwahara, 1970). These findings suggest a suppressor evolution against wCauA-induced MK in E. cautella. In addition, Wolbachia typically encode a number of wmk homologs in their genomes, but their effects differ among insects and are not apparent in some species such as Homona, possibly due to the diversity in sex determination system or the evolution of suppressors. Nonetheless, given the high copy number of wmk homologs in Wolbachia genomes, they may still contribute to Wolbachia biology as transcriptional regulators and potential virulence factors. Virulence genes often undergo duplication and substitution under strong selective pressure (Hill et al., 2022; Jones & Dangl, 2006). An intense evolutionary arms race between Wolbachia and their hosts could have increased and diversified the MK-associated Wolbachia genes. Conversely, natural selection favors the rescue of males by suppressing the Wolbachia-induced reproductive manipulations (Hornett et al., 2006; Hornett et al., 2014), which may involve changes in the sex determination system because Wolbachia strains frequently hijack host reproduction systems. In this context, Wolbachia and other MK-inducing microbes may be a hidden driver for the diversification of complex insect sex determination systems.

In this study, we clarified the conserved roles of the Wolbachia-encoded oscar homologs in Lepidoptera and demonstrated that Wolbachia have evolved distinct MK mechanisms (through causative genes) in insect taxa. The diversification of phenotype/virulence-associated genes and the rampant horizontal transmission of phages carrying virulence genes between Wolbachia strains may have contributed to the outstanding success of this bacterial genus. In addition to oscar and wmk, Wolbachia may retain other uncharacterized genes that induce male lethality, and further studies on diverse Wolbachia–host systems are highly warranted. Our findings provide insights into the molecular mechanisms and evolutionary relationships between endosymbionts and their hosts, which may also contribute to the design of pest management strategies.

Materials and methods

Experimental model and subject details

Homona magnanima

A laboratory-maintained H. magnanima line with a normal sex ratio (NSR) that was negative for Wolbachia and other endosymbionts, was used in our experiments. This line was initially collected in Tokyo, Japan, in 1999 and has been maintained inbred for over 250 generations in the laboratory. Larvae were reared using the artificial SilkMate 2S diet (Nosan Co., Yokohama, Japan) at 25°C under a long photoperiod (16L:8D) (Arai et al., 2019).

The laboratory-maintained all female H. magnanima WT12 line, which was initially collected in Taiwan (Tea Research Extension Station, Taoyuan city) in 2015 (Arai et al., 2020), was also used in this study. This line was maintained for over 50 generations by crossing it with the males of the NSR line.

BmN-4 cell line

BmN-4 cells (provided by Chisa Yasunaga-Aoki, Kyushu University, and maintained in our laboratory) were cultured at 26°C in IPL-41 Insect Medium (Applichem, Darmstadt, Germany) supplemented with 10% fetal bovine serum.

Transient expression of MK-associated phage genes

mRNA synthesis

Codon-optimized wmk genes (wmk-1 to wmk-4), conjugated wmk pairs using a T2A peptide (i.e., wmk-1-T2A-wmk-2 and wmk-3-T2A-wmk-4), and Hm-oscar genes synthesized by Arai et al. (2023b) were used for mRNA synthesis. These synthetic genes were ligated into the plasmid pIZ/V5-His (Invitrogen, MA, USA) using the NEBuilder® HiFi DNA Assembly kit (New England Biolabs, MA, USA) following the manufacturer’s protocol. The inserts of the vector (i.e., wmk-1, wmk-2, wmk-3, wmk-4, Hm-oscar, wmk-1-T2A-wmk-2, wmk-3-T2A-wmk-4, and GFP as a control gene) were amplified using KOD-one (TOYOBO, Osaka, Japan) with the primer set containing the T7 promoter described in Fukui et al. (2015) (i.e., pIZ-F-T7: 5′-TAATACGACTCACTATAGGGAGACAGTTGAACAGCATCTGTTC-3′ and pIZ-R: 5′-GACAATACAAACTAAGATTTAGTCAG-3′) under the following PCR conditions: 20 cycles at 98°C for 10 s, 62°C for 5 s, and 68°C for 15 s. The amplicons were purified using the QIAquick PCR purification kit (Qiagen, Hilden, Germany), and 100–200 ng DNA was used for mRNA synthesis. The capped mRNA (cRNA) with poly(A) tail was synthesized using the mMESSAGE mMACHINE® T7 Ultra kit (Invitrogen, MA, USA) following the manufacturer’s protocol with some modifications. In brief, the assembled transcription reaction (10 µL of T7 2X NTP/ARCA, 2 µL of 10X T7 reaction buffer, 2 µL of T7 enzyme mix, and 6 µL of linear DNA template diluted with nuclease-free water) was incubated at 37°C for at least 10 h to maximize RNA yields. After poly(A) tailing, cRNA was purified using ISOGEN II (Nippongene, Tokyo, Japan) and dissolved in up to 10 µL of nuclease-free water to achieve an RNA concentration of approximately 1500–2000 ng/µL. The synthesized cRNA was preserved at −80°C until further use.

Preparation of H. magnanima eggs

A total of 15 males and 10 females of H. magnanima were mated in a plastic box (30 cm × 20 cm × 5 cm) for 3–4 days. The collection of egg masses began the day after oviposition was confirmed. In brief, newly oviposited egg masses were collected at 30-min intervals during the dark period using red light (which the moths cannot perceive). Females started to oviposit eggs at least 4 h into the dark period and, within less than 30 min post oviposition (mpo), the egg masses were collected. The collection lasted until the start of the light period. The egg masses were then subjected to microinjection assays.

Inoculum preparation, microinjection, and maintenance of the injected embryos

A glass needle for microinjection was prepared using glass capillary GD-1 (Narishige, Tokyo, Japan) with a PC-10 puller (Narishige). The glass capillaries were pulled at two temperatures (first stage: 75°C, second stage: 65°C) using two heavy weights and one light weight. The movement position during the second heating stage was set to 3 mm (range 1–10 mm).

The cRNA solution was diluted in a buffer (100 mM potassium acetate, 2 mM magnesium acetate, and 30 mM HEPES-KOH; at pH 7.4) containing 0.2% (W/V) Brilliant Blue FCF (Wako, Osaka, Japan) to obtain an RNA concentration of 1000 ng/µL. Approximately 1–4 µL of dye-containing mRNA solution was aspirated into the glass needles (capillaries), and the edge of each needle was ground using Micro Grinder EG-402 (Narishige) at an angle of 20 degrees for 2 s.

The fresh egg masses of H. magnanima (collected at less than 30 mpo as described above) were put on a double-sided sticky tape (15 mm × 5 m T4612, Nitoms, Tokyo, Japan) and placed on glass slides. Under a Nikon SMZ1270 microscope (Nicon, Tokyo, Japan), the RNA solution (30–100 fL) was injected into individual eggs using an microinjector IM-400 (Narishige) (balance pressure set at 0.010–0.030; injection pressure set at 0.050-0.100) and a QP-3JOY-2R electric micromanipulator (MicroSupport, Shizuoka, Shizuoka).

The injected egg masses were maintained in a 90-mm plastic Petri dish fitted with a slightly wet filter paper. As high humidity interfered with the development of the embryos, the injected eggs (egg mass) were first maintained at a relative humidity of 30%–50% (0–3 days post injection) and then at a higher humidity until hatching (4–6 dpo, 60% relative humidity). The hatched larvae were reared separately with 1/2 ounces of SilkMate 2S (Nosan Co.) until eclosion. The adult H. magnanima moths that emerged from the cRNA-injected embryos were sexed based on their external morphology.

Sexing of embryos/neonates and RNA extraction

To verify the effect of the transient expression of Hm-oscar on sex determination in H. magnanima, Hm-oscar/GFP-expressed, non-injected, and wHm-t-infected mature embryos showing black head capsule (1 day before hatching, 5–6 days post oviposition, dpo) were dissected on glass slides using forceps, as described in Arai et al. (2022). Their Malpighian tubules were fixed with methanol/acetic acid (50% v/v) and stained with lactic acetic orcein for W chromatin observations. The remaining tissues not used for sexing were stored in ISOGEN II (Nippon Gene) at −80°C until subsequent extraction. In total, 12 males or females (confirmed based on the presence or absence of W chromatin) were pooled and homogenized in ISOGEN II to extract RNA as described in Arai et al. (2023a). In brief, 0.4 times the volume of UltraPure distilled water (Invitrogen) was added to the ISOGEN II homogenates, which were centrifuged at 12000 × g for 15 min at 4°C to pellet proteins and DNAs. The resulting supernatant was mixed with the same volume of isopropanol to precipitate RNAs; then, the resulting solutions were transferred to EconoSpin columns (Epoch Life Science) and centrifuged at 17,900 × g and 4°C for 2 min. The RNAs captured in the column were washed twice with 80% ethanol and eluted in 20 µL of UltraPure distilled water (Invitrogen).

Hmdsx detection

Sex-specific dsx splicing variants of H. magnanima were detected as described in Arai et al. (2023a). In brief, total RNA (100–300 ng) extracted from sex-determined mature embryos was reverse-transcribed using PrimeScript™ II Reverse Transcriptase (TaKaRa, Shiga, Japan) at 30°C for 10 min, 45°C for 60 min, and 70°C for 15 min. Then, cDNA was amplified using KOD-FX Neo (Toyobo Co., Ltd.) with the following two primers: Hmdsx_long3F (5′-TGCCTAAAGTGAAAACGCCGAGGAGCC-3′) and Hmdsx_Mrev (5′-TGGAGGTCTCTTTTCATCCGG-3′). The PCR conditions used were as follows: 94°C for 2 min, followed by 45 cycles of 98°C for 10 s, 66°C for 30 s, and 68°C for 30 s. The amplicons were subjected to electrophoresis on 2.0% agarose Tris-borate-EDTA buffer (89 mM Tris-borate, 89 mM boric acid, 2 mM EDTA) gels.

Quantification of Z chromosome-linked genes

The effects on dosage compensation were assessed by measuring differences in gene expression, as described in Fukui et al. (2015) and Arai et al. (2023a). A total of 1.0 µg of the total RNA extracted from Hm-oscar or GFP-overexpressed mature embryos (108 hpo, approximately 40-50 sex determined embryos) was used to prepare mRNA-seq libraries via the NEBNext Poly (A) mRNA Magnetic Isolation Module (New England Biolabs) and the NEBNext Ultra II RNA Library Prep kit for Illumina (New England Biolabs) following the manufacturer’s protocol. The adaptor sequences and low-quality reads (Qscore <20) were removed from the generated sequence data [150 bp paired-end (PE150)] using Trimmomatic (Bolger et al., 2014). The trimmed reads were mapped to the previously assembled transcriptome database for H. magnanima (Arai et al., 2023a) using Kallisto (Bray et al., 2016) to generate the normalized read count data (transcripts per million, TPM). While we did not generate new RNA-seq data of the wHm-t-infected and uninfected groups, the expression data included in Arai et al. (2023a) were re-analyzed along with the newly obtained RNA-seq data above. The binary logarithms of the TPM differences between males and females belonging to each H. magnanima group (i.e., Hm-oscar/GFP-expressed) were calculated to assess the fold-changes in gene expression levels between sexes. As described in Arai et al. (2023a), the transcriptome data of H. magnanima were annotated using the B. mori gene sets obtained from KAIKOBASE (https://kaikobase.dna.affrc.go.jp). The binary logarithms of the TPM differences between males and females in the B. mori chromosomes 1 to 28 were plotted.

To confirm the effects observed in the RNA-seq-based quantification of the Z chromosome genes, expression level of the Z-linked Hmtpi gene was quantified by reverse transcription-qPCR as described in Arai et al. (2023a). Briefly, total RNA (100–300 ng) extracted from pooled male or female mature embryos (108 hpo) was reverse transcribed using PrimeScript II reverse transcriptase (TaKaRa) as described above. The cDNA was used to quantify relative gene expression levels, with normalization to the control gene elongation factor 1a (ef1a). qPCR was performed using primer sets for Hmtpi (HmTpi_F180702_212: 5′-GCTGCGAGTGGGTGATTTTG-3′; HmTpi_R180702_326: 5′-GCGATCACTTTCAGACCCGA-3′) and Hmef1a (Hmef1a_F_val1_85: 5′-TTTCCAGGGTGGTTGAGCA-3′; Hmef1a_R_val1_193: 5′-CCGTTAAGGAGCTGCGTCG-3′), and KOD SYBR qPCR Mix (Toyobo, Osaka, Japan) in a LightCycler 96 system (Roche, Basel, Switzerland). The qPCR reaction was made with 5 µl KOD SYBR (TOYOBO), 0.4 µL forward primer [10 pmol/µl], 0.4 µl reverse primer [10 pmol/µl], and 3.2 µl water. The qPCR was performed under the following conditions: 180 s at 95°C, 40 cycles of 8 s at 98°C, 10 s at 60°C, and 10 s at 68°C, followed by heating to 90°C for melting curve analysis. Mean cycle threshold (CT) values of samples were calculated for at least 13 replicates, and both ΔCT (CT Z gene – CT ef1a) and ΔΔCT (ΔCTmale – ΔCTfem) values were calculated.

Transfection assays and quantification of BmIMPM

The coding sequence of Hmmasc was amplified from cDNA derived from the RNA extracted from male embryos of the NSR line using KOD-one (TOYOBO) with the following primer set: HmMasc_CDS_HindIIIF: 5′-GCAAAGCTTCAACATGATCTCTCGCCAACCACAATCAACATCA-3′ and HmMasc_CDS_BamHI R: 5′-GCAGGATCCCAACCTACTGATAAGGAGGGAAGTAAGGCTGCTG-3′. The following PCR conditions were applied: 20 cycles of 98°C for 10 s, 62°C for 5 s, and 68°C for 15 s. The amplicon was purified with the QIAquick PCR purification kit (Qiagen) and cloned into the pIZ/V5-His vector using the HindIII and BamHI enzymes (New England Biolabs). The codon-optimized oscar gene of wFur (Katsuma et al. 2022) was also cloned into the pIZ/V5-His vector using the KpnI and NotI enzymes (TaKaRa).

To verify the masculinizing function of Hmmasc, BmN-4 cells (4 × 105 cells per dish, diameter 35 mm) were transfected with 1 µg of each plasmid DNA (pIZ/V5-His having Hmmasc) using FuGENE HD (Promega, WI, USA), as described in Katsuma et al. (2022). To clarify whether Hm-oscar suppressed the function of Hmmasc, 1 µg of plasmid DNA (pIZ/V5-His bearing either Hmmasc or Hm-oscar and non-inserted vector as negative control) was transfected to the BmN-4 cells using FuGENE HD (Promega). Three days after transfection, the cells were collected and subjected to RNA extraction via TRI REAGENT® (Molecular Research Center Inc., USA) and cDNA construction with AMV transcriptase (TaKaRa). The degree of masculinization in the BmN-4 cells (default female-type sex determination) was verified by quantifying the expression levels of BmImpM (Suzuki et al., 2010), which is involved in male-specific sex determination cascades, using primers BmIMP_F: 5′-ATGCGGGAAGAAGGTTTTATG-3′ and BmIMP_R: 5′-TCATCCCGCCTCAGACGATTG-3′, as described in Fukui et al. (2015). BmRp49 gene was also amplified as a control gene using primers rp49_F: 5′-CCCAACATTGGTTACGGTTC-3′ and rp49_R: 5′-GCTCTTTCCACGATCAGCTT-3′. Further interactions between Hm-oscar and masc genes derived from lepidopteran insects [i.e., Trilocha varians masc (Tvmasc), Spodoptera frugiperda masc (Sfmasc), B. mori masc (Bmmasc), O. furnacalis masc (Ofmasc), and Papilio machaon masc (Pmmasc)] (Katsuma et al., 2022b) were assessed using the same procedures. The expression levels of BmImpM were analyzed as described in Katsuma et al. (2022). In brief, the relative expression of BmImpM (CT BmImpM / CT BmRp49) was calculated for each experimental group. Then, relative expression of BmImpM of each sample was normalized by setting the mean in the Hmmasc singly transfected group as 100. The normalized relative expression of BmImpM was visualized using R v4.0 with ggplot2 (https://ggplot2.tidyverse.org/).

Quantification and statistical analysis

The number of surviving H. magnanima injected with either Hm-oscar, wmks, or GFP in each replicate were counted at the adult stage. The male ratios (number of adult males/numbers of all adults) under all conditions were compared using the Steel– Dwass test in R v4.0. The sex ratio bias was also assessed based on the total numbers of (i) male and female adults and (ii) male and female embryos under each condition using the binomial test in R v4.0.

Gene expression levels between chromosomes (RNA-seq), a Z-linked Hmtpi gene expression in H. magnanima (RT-qPCR), and normalized BmIMPM expression in BmN-4 cells under all conditions (RT-qPCR) were compared using the Steel– Dwass test in R v4.0.

A phylogenetic tree of Masc proteins used in this study was constructed based on maximum likelihood with bootstrap re-sampling of 1,000 replicates using MEGA7 (Kumar et al., 2016).

eLife Assessment

This valuable study implicates a specific Wolbachia gene in driving the male-killing phenotype in a moth: This is a contribution to a growing body of literature from the authors in which they authors have nicely teased apart the loci responsible for male killing across diverse insects. The conclusions are supported by solid evidence.

Reviewer #1 (Public review):

Summary:

Insects and their relatives are commonly infected with microbes that are transmitted from mothers to their offspring. A number of these microbes have independently evolved the ability to kill the sons of infected females very early in their development; this male killing strategy has evolved because males are transmission dead-ends for the microbe. A major question in the field has been to identify the genes that cause male killing and to understand how they work. This has been especially challenging because most male-killing microbes cannot be genetically manipulated. This study focuses on a male-killing bacterium called Wolbachia. Different Wolbachia strains kill male embryos in beetles, flies, moths, and other arthropods. This is remarkable because how sex is determined differs widely in these hosts. Two Wolbachia genes have been previously implicated in male-killing by Wolbachia: oscar (in moth male-killing) and wmk (in fly male-killing). The genomes of some male-killing Wolbachia contain both of these genes, so it is a challenge to disentangle the two.

This paper provides strong evidence that oscar is responsible for male-killing in moths. Here, the authors study a strain of Wolbachia that kills males in a pest of tea, Homona magnanima. Overexpressing oscar, but not wmk, kills male moth embryos. This is because oscar interferes with masculinizer, the master gene that controls sex determination in moths and butterflies. Interfering with the masculinizer gene in this way leads the (male) embryo down a path of female development, which causes problems in regulating the expression of genes that are found on the sex chromosomes.

Strengths:

The authors use a broad number of approaches to implicate oscar, and to dissect its mechanism of male lethality. These approaches include: a) overexpressing oscar (and wmk) by injecting RNA into moth eggs, b) determining the sex of embryos by staining female sex chromosomes, c) determining the consequences of oscar expression by assaying sex-specific splice variants of doublesex, a key sex determination gene, and by quantifying gene expression and dosage of sex chromosomes, using RNASeq, and d) expressing oscar along with masculinizer from various moth and butterfly species, in a silkmoth cell line. This extends recently published studies implicating oscar in male-killing by Wolbachia in Ostrinia corn borer moths, although the Homona and Ostrinia oscar proteins are quite divergent. Combined with other studies, there is now broad support for oscar as the male-killing gene in moths and butterflies (i.e. order Lepidoptera). So an outstanding question is to understand the role of wmk. Is it the master male-killing gene in insects other than Lepidoptera and if so, how does it operate?

Weaknesses:

I found the transfection assays of oscar and masculinizer in the silkworm cell line (Figure 4) to be difficult to follow. There are also places in the text where more explanation would be helpful for non-experts.

Reviewer #2 (Public review):

Summary:

Wolbachia are maternally transmitted bacteria that can manipulate host reproduction in various ways. Some Wolbachia induce male killing (MK), where the sons of infected mothers are killed during development. Several MK-associated genes have been identified in Homona magnanima, including Hm-oscar and wmk-1-4, but the mechanistic links between these Wolbachia genes and MK in the native host are still unclear.

In this manuscript, Arai et al. show that Hm-oscar is the gene responsible for Wolbachia-induced MK in Homona magnanima. They provide evidence that Hm-Oscar functions through interactions with the sex determination system. They also found that Hm-Oscar disrupts sex determination in male embryos by inducing female-type dsx splicing and impairing dosage compensation. Additionally, Hm-Oscar suppresses the function of Masc. The manuscript is well-written and presents intriguing findings. The results support their conclusions regarding the diversity and commonality of MK mechanisms, contributing to our understanding of the mechanisms and evolutionary aspects of Wolbachia-induced MK.

Comments on revisions:

The authors have already addressed the reviewer's concerns.

Reviewer #3 (Public review):

Summary:

Overall, this is a clearly written manuscript with nice hypothesis testing in a non-model organism that addresses the mechanism of Wolbachia-mediated male killing. The authors aim to determine how five previously identified male-killing genes (encoded in the prophage region of the wHm Wolbachia strain) impact the native host, Homona magnanima moths. This work builds on the authors' previous studies in which

(1) they tested the impact of these same wHm genes via heterologous expression in Drosophila melanogaster

(2) also examined the activity of other male-killing genes (e.g., from the wFur Wolbachia strain in its native host: Ostrinia furnacalis moths).

Advances here include identifying which wHm gene most strongly recapitulates the male-killing phenotype in the native host (rather than in Drosophila), and the finding that the Hm-Oscar protein has the potential for male-killing in a diverse set of lepidopterans, as inferred by the cell-culture assays.

Strengths:

Strengths of the manuscript include the reverse genetics approaches to dissect the impact of specific male-killing loci, and use of a "masculinization" assay in Lepidopteran cell lines to determine the impact of interactions between specific masc and oscar homologs.

Weaknesses:

It is clear from Figure 1 that the combinations of wmk homologs do not cause male killing on their own here. While I largely agree with the author's conclusions that oscar is the primary MK factor in this system, I don't think we can yet rule out that wmk(s) may work synergistically or interactively with oscar in vivo. This might be worth a small note in the discussion. (eg at line 294 'indicating that wmk likely targets factors other than masc." - this could be downstream of the impacts of oscar; perhaps dependent on oscar-mediated impacts on masc first).

Regarding the perceived male-bias in Figure 2a: I think readers might be interpreting "unhatched" as "total before hatching". You could eliminate ambiguity by perhaps splitting the bars into male and female, and then within a bar, coloring by hatched versus unhatched. But this is a minor point, and I think the updated text helps clarify this.

The new Figure 4b looks to be largely redundant with the oscar information in Figure 1a.

Updated statistical comparisons for the RNA-seq analysis are helpful. However these analyses are based on single libraries (albeit each a pool of many individuals), so this is still a weaker aspect of the manuscript.

The new information on masc similarity is useful (Fig 4d) - if the authors could please include a heatmap legend for the colors, that would be helpful. Also, please avoid green and red in the same figure when key for interpretation.

Figure 1A "helix-turn-helix" is misspelled. ("tern").

Author response:

The following is the authors’ response to the previous reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

Insects and their relatives are commonly infected with microbes that are transmitted from mothers to their offspring. A number of these microbes have independently evolved the ability to kill the sons of infected females very early in their development; this male killing strategy has evolved because males are transmission dead-ends for the microbe. A major question in the field has been to identify the genes that cause male killing and to understand how they work. This has been especially challenging because most male-killing microbes cannot be genetically manipulated. This study focuses on a male-killing bacterium called Wolbachia. Different Wolbachia strains kill male embryos in beetles, flies, moths, and other arthropods. This is remarkable because how sex is determined differs widely in these hosts. Two Wolbachia genes have been previously implicated in male-killing by Wolbachia: oscar (in moth male-killing) and wmk (in fly male-killing). The genomes of some male-killing Wolbachia contain both of these genes, so it is a challenge to disentangle the two.

This paper provides strong evidence that oscar is responsible for male-killing in moths. Here, the authors study a strain of Wolbachia that kills males in a pest of tea, Homona magnanima. Overexpressing oscar, but not wmk, kills male moth embryos. This is because oscar interferes with masculinizer, the master gene that controls sex determination in moths and butterflies. Interfering with the masculinizer gene in this way leads the (male) embryo down a path of female development, which causes problems in regulating the expression of genes that are found on the sex chromosomes.

We would like to thank you for evaluating our manuscript.

Strengths:

The authors use a broad number of approaches to implicate oscar, and to dissect its mechanism of male lethality. These approaches include: a) overexpressing oscar (and wmk) by injecting RNA into moth eggs, b) determining the sex of embryos by staining female sex chromosomes, c) determining the consequences of oscar expression by assaying sex-specific splice variants of doublesex, a key sex determination gene, and by quantifying gene expression and dosage of sex chromosomes, using RNASeq, and d) expressing oscar along with masculinizer from various moth and butterfly species, in a silkmoth cell line. This extends recently published studies implicating oscar in male-killing by Wolbachia in Ostrinia corn borer moths, although the Homona and Ostrinia oscar proteins are quite divergent. Combined with other studies, there is now broad support for oscar as the male-killing gene in moths and butterflies (i.e. order Lepidoptera). So an outstanding question is to understand the role of wmk. Is it the master male-killing gene in insects other than Lepidoptera and if so, how does it operate?

We would like to thank you for evaluating our manuscript. Our data demonstrated that Oscar homologs play important roles in male-killing phenotypes in moths and butterflies; however, the functional relevance of wmk remains uncertain. As you noted, whether wmk acts as a male-killing gene in insects such as flies and beetles—or even in certain lepidopteran species—requires further investigation using diverse insect models, which we are eager to explore in future research.

Weaknesses:

I found the transfection assays of oscar and masculinizer in the silkworm cell line (Figure 4) to be difficult to follow. There are also places in the text where more explanation would be helpful for non-experts.

Thank you for your suggestion. We have revised the section on the cell-based experiment. Further, we revised the manuscript to make it accessible to a broader audience. We believe these revisions have significantly improved the clarity and comprehensiveness of our manuscript.

Reviewer #2 (Public review):

Summary:

Wolbachia are maternally transmitted bacteria that can manipulate host reproduction in various ways. Some Wolbachia induce male killing (MK), where the sons of infected mothers are killed during development. Several MK-associated genes have been identified in Homona magnanima, including Hm-oscar and wmk-1-4, but the mechanistic links between these Wolbachia genes and MK in the native host are still unclear.

In this manuscript, Arai et al. show that Hm-oscar is the gene responsible for Wolbachia-induced MK in Homona magnanima. They provide evidence that Hm-Oscar functions through interactions with the sex determination system. They also found that Hm-Oscar disrupts sex determination in male embryos by inducing female-type dsx splicing and impairing dosage compensation. Additionally, Hm-Oscar suppresses the function of Masc. The manuscript is well-written and presents intriguing findings. The results support their conclusions regarding the diversity and commonality of MK mechanisms, contributing to our understanding of the mechanisms and evolutionary aspects of Wolbachia-induced MK.

We would like to thank you for evaluating our manuscript.

Comments on revisions:

The authors have already addressed the reviewer's concerns.

We would like to thank you for evaluating our manuscript.

Reviewer #3 (Public review):

Summary:

Overall, this is a clearly written manuscript with nice hypothesis testing in a non-model organism that addresses the mechanism of Wolbachia-mediated male killing. The authors aim to determine how five previously identified male-killing genes (encoded in the prophage region of the wHm Wolbachia strain) impact the native host, Homona magnanima moths. This work builds on the authors' previous studies in which

(1) they tested the impact of these same wHm genes via heterologous expression in Drosophila melanogaster

(2) also examined the activity of other male-killing genes (e.g., from the wFur Wolbachia strain in its native host: Ostrinia furnacalis moths).

Advances here include identifying which wHm gene most strongly recapitulates the male-killing phenotype in the native host (rather than in Drosophila), and the finding that the Hm-Oscar protein has the potential for male-killing in a diverse set of lepidopterans, as inferred by the cell-culture assays.

We would like to thank you for evaluating our manuscript.

Strengths:

Strengths of the manuscript include the reverse genetics approaches to dissect the impact of specific male-killing loci, and use of a "masculinization" assay in Lepidopteran cell lines to determine the impact of interactions between specific masc and oscar homologs.

We would like to thank you for evaluating our manuscript.

Weaknesses:

It is clear from Figure 1 that the combinations of wmk homologs do not cause male killing on their own here. While I largely agree with the author's conclusions that oscar is the primary MK factor in this system, I don't think we can yet rule out that wmk(s) may work synergistically or interactively with oscar in vivo. This might be worth a small note in the discussion. (eg at line 294 'indicating that wmk likely targets factors other than masc." - this could be downstream of the impacts of oscar; perhaps dependent on oscar-mediated impacts on masc first).

We sincerely appreciate your suggestion. Whilst wmk genes themselves did not exhibit apparent lethal effects on the native host, as you noted, we cannot entirely rule out the possibility that wmk may be involved in male-killing actions, either directly or indirectly assisting the function of Hb-oscar. Following your suggestion, we have added a brief note in the discussion section regarding the interpretation of wmk functions.

“In addition, Katsuma et al. (2022) reported that the wmk homologs encoded by wFur did not affect the masculinizing function of masc in vitro, indicating that wmk likely targets factors other than masc. Whilst we cannot rule out the possibility that wmk may work synergistically or interactively with oscar in vivo—potentially acting downstream of oscar’s impact—our results strongly suggested that Wolbachia strains have acquired multiple MK genes through evolution.” (lines 287-292)

Regarding the perceived male-bias in Figure 2a: I think readers might be interpreting "unhatched" as "total before hatching". You could eliminate ambiguity by perhaps splitting the bars into male and female, and then within a bar, coloring by hatched versus unhatched. But this is a minor point, and I think the updated text helps clarify this.

Thank you for your suggestion. We have accordingly revised the figure 2a. In addition, we have included more detailed information in the first sentence of the section Males are killed mainly at the embryonic stage.

“The sex of hatched larvae (neonates) and the remaining unhatched embryos was determined by the presence or absence of W chromatin, a condensed structure of the female-specific W chromosome observed during interphase.” (lines 171-173)

The new Figure 4b looks to be largely redundant with the oscar information in Figure 1a.

Thank you for your suggestion. We have removed Figure 4b due to its overlap with Figure 1a and have incorporated relevant figure legends into the Figure 1a legend.

Updated statistical comparisons for the RNA-seq analysis are helpful. However these analyses are based on single libraries (albeit each a pool of many individuals), so this is still a weaker aspect of the manuscript.

Thank you for your suggestion. As you noted, the use of single libraries (due to the limited number of available individuals, though each includes approximately 50 males and females) may be a potential limitation of this study. However, as demonstrated in the qPCR assay for the Z-linked gene provided in the previous revision, we believe that our data and conclusion—that Wolbachia/ Hb-oscar disrupts dosage compensation by causing the overexpression of Z-linked genes—are well-supported and robust.

The new information on masc similarity is useful (Fig 4d) - if the authors could please include a heatmap legend for the colors, that would be helpful. Also, please avoid green and red in the same figure when key for interpretation.

Thank you for your suggestion. We have accordingly included a heatmap legend and revised the colors.

Figure 1A "helix-turn-helix" is misspelled. ("tern").

We have revised.

Recommendations for the authors:

Comments from the reviewing editor: I would suggest you address the comments of the reviewer on the revised version.

We have further revised the manuscript to address all the questions, comments and suggestions provided by the reviewers. We believe that the resulting revisions have significantly enhanced the quality and comprehensiveness of our manuscript.

Reviewer #1 (Recommendations for the authors):

Thank you for revising this manuscript. I have a few last recommendations:

- Line 214: re: 'Statistical data are available in the supplementary data file', it would be more helpful to add a few words here that actually summarize the statistical results

We would like to thank you for your suggestion. We have revised the sentence to describe the overview of the statistical results.

“RNA-seq analysis revealed that, in Hm-oscar-injected embryos, Z-linked genes (homologs on the B. mori chromosomes 1 and 15) were more expressed in males than in females (Fig. 3a), which was not observed in the GFP-injected group (Fig. 3b). Similarly, as previously reported by Arai et al. (2023a), high levels of Z-linked gene expression were also observed in wHm-t-infected males, but not in NSR males (Fig. 3c,d). The high (i.e., doubled) Z-linked gene expression in both Hm-oscar-expressed and wHm-t-infected males was further confirmed by quantification of the Z-linked Hmtpi gene (Fig. 3e). These trends were statistically supported, with all data available in the supplementary data file.” (lines 205-213)

- Figure 1 legend: do you mean 'bridged' instead of 'brigged'?

We have accordingly revise, thank you for the suggestion.

Version 3

Abstract

Abstract

Wolbachia are maternally transmitted bacterial symbionts that are ubiquitous among arthropods. They can hijack host reproduction in various ways, including male killing (MK), where the sons of infected mothers are killed during development. The recent discovery of MK-associated Wolbachia genes, i.e., oscar in Ostrinia moths and wmk in Drosophila flies, stimulates our interest in the diversity and commonality of MK mechanisms, which remain largely unclear. We recently discovered that a Wolbachia symbiont of the moth Homona magnanima carries an MK-associated prophage region encoding homologs of oscar (Hm-oscar) and wmk (wmk-1–4). Here, we investigated the effects of these genes in the native host. Upon transient overexpression, Hm-oscar, but not wmk, induced male lethality in H. magnanima, in contrast to our observations in Drosophila, where the wmk homologs, but not Hm-oscar, killed the males. Hm-oscar disrupted sex determination in male embryos by inducing a female-type doublesex splicing and impaired dosage compensation, recapitulating the Wolbachia phenotype. Cell-based transfection assays confirmed that Hm-oscar suppressed the function of masculinizer, the primary male sex determinant involved in lepidopteran dosage compensation. Our study highlights the conserved roles of oscar homologs in Wolbachia-induced lepidopteran MK and argues that Wolbachia have evolved multiple MK mechanisms in insects.

Introduction

Arthropods commonly carry microbial symbionts that are passed from mother to offspring (Hurst & Frost, 2015; Werren et al., 2008; Hurst, 2017). The maternally transmitted bacteria, belonging to the genus Wolbachia (Alphaproteobacteria), are estimated to be present in at least 40% of all insect species, making them one of the most widespread endosymbiont genera on the planet (Werren et al, 2008; Zug & Hammerstein, 2012). Wolbachia have achieved evolutionary success by manipulating host reproduction through various means that enhance endosymbiont transmission (Werren et al., 2008). Such manipulation of the host’s reproduction includes cytoplasmic incompatibility (CI), parthenogenesis, feminization, and male killing (MK), each of which affects the biological features, distribution, and evolution of the host. Among these strategies, MK directly skews the sex ratio of the host population toward females by killing male offspring of infected mothers during development. The lack of symbiont transmission through male hosts often leads to the evolution of MK. Wolbachia have been shown to induce MK in a wide range of insect taxa (e.g., dipterans, lepidopterans, and coleopterans) and other arthropods (Hurst et al., 1999; Kageyama & Traut, 2004; Hurst et al., 2000). Furthermore, various bacteria, viruses, and microsporidia induce MK phenotypes (Kageyama et al., 2012; Fujita et al., 2021; Kageyama et al., 2023; Nagamine et al., 2023), and recent studies have postulated that these microbes have evolved their MK ability independently (Harumoto & Lemaitre, 2018; Kageyama et al., 2023; Nagamine et al., 2023; Arai et al., 2023a).

The evolution and molecular mechanisms underlying Wolbachia-induced MK, observed across diverse insect taxa, have attracted considerable attention for decades. MK mechanisms are hypothesized to be linked to sex determination cascades regulating male and female differentiation in insects (Hornett et al., 2022). Indeed, some MK-inducing Wolbachia strains disrupt the sex determination system in host moths (Sugimoto and Ishikawa, 2012; Sugimoto et al., 2015; Fukui et al., 2015; Arai et al., 2023a). Lepidopteran insects (moths and butterflies) generally have a female heterogametic sex chromosome system (e.g., WZ in females, ZZ in males) and employ dosage compensation, which equalizes the Z-chromosome-linked gene dose between sexes (Kiuchi et al., 2014). Dosage compensation is regulated by the masculinizer gene (masc), which is critical for male development. The lepidopteran sex determination system consists of multiple transcriptional regulators, some of which exhibit sex-linked expression and/or splicing isoforms. The primary male sex determinant, masc, also regulates the downstream master sex determination gene, doublesex (dsx), which exhibits sex-dependent splicing isoforms (dsxF in females and dsxM in males) (Kiuchi et al., 2014). In Ostrinia and Homona male moths, MK-inducing Wolbachia strains impair dosage compensation by disrupting the expression of masc. Furthermore, they disrupt sex determination in male moths by inducing the “female” isoform of dsx (dsxF), leading to a “mismatch” between genetic sex (male: ZZ sex chromosome constitution) and phenotypic sex (female: based on dsxF), ultimately resulting in male death (Arai et al., 2023a; Fukui et al., 2015; Sugimoto et al., 2015; Sugimoto & Ishikawa, 2012).

More recently, a Wolbachia protein named Oscar (Osugoroshi protein containing CifB C-terminus-like domain and many Ankyrin Repeats; Osugoroshi translates to male killing in Japanese) was shown to recapitulate the Wolbachia wFur-induced MK phenotype in the native host moth Ostrinia furnacalis (Katsuma et al., 2022). The Oscar protein interacts with and degrades Masc protein, leading to the failure of dosage compensation and the production of female-type dsx isoforms in Ostrinia male moths (Fukui et al., 2024; Katsuma et al., 2022). Although oscar homologs are common in MK Wolbachia strains from Lepidoptera, MK strains from Diptera (Drosophila) and from several lepidopteran hosts lacked the gene (Arai et al., 2023b; Arai et al., 2024a). Furthermore, oscar homologs are evolutionarily dynamic, with highly variable sequences and structures (Arai et al., 2024a), making it difficult to assess their functional relevance. In addition to oscar, the helix-turn-helix domain-containing putative transcriptional regulator wmk is widely conserved among Wolbachia strains and induces various toxicities in Drosophila flies, ranging from no effect, weak male lethality (30% mortality), strong male lethality (90%–100% mortality), or complete lethality of both sexes (Perlmutter et al., 2019; Perlmutter et al., 2020; Perlmutter et al., 2021; Arai et al., 2023b). Although the mechanisms underlying the wmk-induced toxicities and their connection to sex determination systems remain unclear, these findings suggest that Wolbachia strains carry multiple factors that cause male lethality. However, the diversity and commonality of these functions in insects remain largely unknown, partly due to the technical challenges in validating gene functions in non-model insects.

We recently discovered a prophage region responsible for a Wolbachia-induced MK by comparing closely related non-MK Wolbachia (wHm-c) and MK Wolbachia (wHm-t) in the tea tortrix moth Homona magnanima (Tortricidae) (Arai et al., 2023b; Arai et al., 2024b). The MK-associated prophage element encodes potential MK causative genes–four wmk homologs (wmk-1 to wmk-4) as well as an oscar homolog (Hm-oscar, encoding 1181 aa protein), which differs significantly in sequence length and structure from the wFur-encoded oscar (encoding 1830 aa protein) (Fig. 1a). When these MK candidate genes are transgenically overexpressed in Drosophila, wmk-1 and wmk-3 have a lethality of almost 100%, while wmk-2, wmk-4, and Hm-oscar induce no lethal effects when singly overexpressed. Furthermore, co-expression of the adjacent wmk-3 and wmk-4 has been shown to induce the death of 90% of male flies and restores female survival, suggesting their combined action (Arai et al., 2023b). However, transgenic expression of genes from a moth-derived male killer in flies does not necessarily replicate the native host–bacteria interaction, and the mechanistic links between prophage-encoded Wolbachia genes and MK in the native host Homona remain unclear.

Transient expressions of the Hm-oscar gene resulted in female-biased sex ratios

(a) Structures of wFur-encoded Oscar and wHm-t-encoded Hm-Oscar and four Wmk proteins. (b) Male ratio of adult progeny obtained from cRNA-injected groups (wmk-1, wmk-2, T2A-bridged wmk-1 and wmk-2, wmk-3, wmk-4, T2A-bridged wmk-3 and wmk-4, GFP, and Hm-oscar; n = 5–15 independent replicates using different egg masses), the non-injected NSR line, and the MK wHm-t-positive WT12 line. The total numbers of adult females and males are shown at the bottom. Different letters indicate significant differences (Steel–Dwass test, P < 0.05). The dot plots show all data points individually.

In this study, we showed that the prophage-encoded Hm-oscar recapitulates Wolbachia-induced MK in H. magnanima. Furthermore, we clarified the mechanistic links to host sex-determination cascades both in vivo and in vitro and discussed the underlying mechanisms of MK in Lepidoptera, arguing for the diverse evolutionary origin of Wolbachia-induced MK.

Results

Hm-oscar induces female-biased sex ratios

To achieve the transient overexpression of Hm-oscar and the four wmk genes (wmk-1, wmk-2, wmk-3, and wmk-4), constructed mRNA (cRNA) was injected into Wolbachia-free H. magnanima embryos. Subsequently, the adult moths that emerged from the cRNA-injected embryos were sexed based on their external morphology. When Hm-oscar was overexpressed, the sex ratio of adults was strongly female-biased (42 males and 218 females in total, 14.4% ± 11.1% male ratio in 15 replicates), which was in sharp contrast with the ratios observed in the GFP-injected (126 females and 123 males in total, 49.1% ± 2.08% male ratio in 10 replicates) and non-injected (NSR) (201 females and 215 males in total, 51.4% ± 3.08% male ratio in 9 replicates) groups (P = 0.001 and 0.002, respectively, Steel–Dwass test, Fig. 1b). Compared with the GFP-injected group, the sex ratio was not biased by the overexpression of wmk-1 (51.1% ± 7.27% male ratio in 6 replicates, P = 0.999), wmk-2 (53.2% ± 4.77% male ratio in 6 replicates, P = 0.810), wmk-3 (53.3% ± 7.91% male ratio in 5 replicates, P = 0.999), and wmk-4 (52.9% ± 3.47% male ratio in 5 replicates, P = 0.773). Although the dual expression of wmk-3 and wmk-4 induces strong male lethality in Drosophila (Arai et al., 2023b), here, it did not skew the sex ratio of Homona compared with the ratio detected in the GFP-injected group (57.9% ± 7.67% male ratio in 5 replicates, P = 0.704). Similarly, the dual expression of the tandemly arrayed wmk-1 and wmk-2 did not bias the sex ratio (49.4% ± 2.82% male ratio in 5 replicates) compared with the value in the GFP-injected group (P = 1.000). The sex ratio data of all replicates are available on the supplementary data file.

Males are killed mainly at the embryonic stage

The sex of hatched larvae (neonates) and the remaining unhatched embryos was determined by the presence or absence of W chromatin, a condensed structure of the female-specific W chromosome observed during interphase. MK Wolbachia wHm-t kills Homona males during embryogenesis, resulting in a female-biased sex ratio in hatched larvae (neonates) and a male-biased sex ratio in unhatched embryos (Arai et al., 2023b; Arai et al., 2020). In line with this finding, the sex ratios of hatched larvae (34 females and 6 males) and unhatched embryos (27 females and 47 males) of the MK wHm-t infected line (WT12) were also female- and male-biased, respectively (P = 0.013 and 0, respectively). In the Hm-oscar-injected group, the sex ratio of the hatched larvae (neonates) was strongly female-biased (21 females and 3 males, P = 0.0001 in the binomial test) (Fig. 2a). In contrast, the unhatched mature embryos showed a male-biased ratio (18 females and 38 males, P = 0.005), suggesting that male mortality primarily occurred during the embryonic stage, reflecting the wHm-t-induced MK phenotype. In the non-injected group (NSR), the sex ratios of hatched larvae (28 females and 28 males) and unhatched embryos (23 females and 25 males) were not biased (P = 0.553 and 0.443, respectively).

Hm-Oscar induced lethality in male embryos with the female-type sex determination

(a) Sex ratio of the hatched larvae and unhatched embryos in the Hm-oscar-expressed, wHm-t-infected, and non-infected/expressed groups. Females and males were discriminated based on the presence or absence of W chromatin. The number of individuals is indicated in brackets. (b) Splicing patterns of the downstream sex-determining gene dsx of H. magnanima embryos (5 days post oviposition). Abbreviations: HmOsc, Hm-oscar injected group; GFP, GFP-injected group; WT12, wHm-t-infected line; NSR, non-infected/injected line. M and F indicate W chromatin-negative (ZZ: male genotype) and W chromatin-positive (ZW: female genotype) mature embryos, respectively. dsx-F and dsx-M represent female and male-specific splicing variants, respectively. L: 100 bp DNA ladder (ExcelBand 100 bp DNA Ladder, SMOBIO Technology, Inc., Hsinchu, Taiwan). 0.1, 0.5, 1.0, and 1.5 kb markers are indicated with arrows. N: negative control (water).

Female-type sex determination in male embryos that are destined to die

Splicing patterns of the downstream sex determinant dsx were assessed in each of the H. magnanima embryos that were Hm-oscar-expressed, GFP-expressed, wHm-t-infected, or non-expressed (i.e., non-injected, NSR). In GFP-expressed and NSR embryos, females and males (which were identified by the presence and absence of W chromatin, respectively) exhibited a female- and male-type splicing variant of dsx, respectively. However, both wHm-t-infected and Hm-oscar-expressed embryos induced female-type dsx splicing regardless of the presence or absence of W chromatin (Fig. 2b). Notably, male embryos expressing Hm-oscar also exhibited weak male-type dsx splicing in addition to the female-type splicing, resembling the previously observed pattern in male embryos infected with low-titer wHm-t (Arai et al., 2023a).

Hm-oscar impairs dosage compensation in male embryos

To confirm the effects of Hm-oscar on dosage compensation, we quantified the expression of Z-linked genes using RNA-seq and qPCR. Unlike other lepidopteran species, the family Tortricidae, including H. magnanima, generally possess a large Z chromosome that is homologous to B. mori chromosomes 1 (Z) and 15 (autosome). RNA-seq analysis revealed that, in Hm-oscar-injected embryos, Z-linked genes (homologs on the B. mori chromosomes 1 and 15) were more expressed in males than in females (Fig. 3a), which was not observed in the GFP-injected group (Fig. 3b). Similarly, as previously reported by Arai et al. (2023a), high levels of Z-linked gene expression were also observed in wHm-t-infected males, but not in NSR males (Fig. 3c,d). The high (i.e., doubled) Z-linked gene expression in both Hm-oscar-expressed and wHm-t-infected males was further confirmed by quantification of the Z-linked Hmtpi gene (Fig. 3e). These trends were statistically supported, with all data available in the supplementary data file.

Hm-oscar-overexpressed male embryos showed higher levels of Z-linked gene expression

(a-d) Normalized expression levels (TPM) and chromosomal distributions of transcripts in H. magnanima embryos. RNA-seq data of embryos (108 hpo) were used to make the following comparisons: Hm-oscar-injected males versus Hm-oscar injected females (a), GFP-injected males versus GFP-injected females (b), WT12 males versus WT12 females (c), and NSR males versus NSR females (d). The chromosome number for each H. magnanima transcript-derived contig was assigned based on Bombyx mori gene models. The X axis represents the chromosome number of B. mori (shown as chr01 to chr28), and chr01 and chr15 (highlighted in orange) correspond to the Z chromosome of H. magnanima (Arai et al., 2023a). (e) Relative expression of a Z-linked tpi gene in H. magnanima groups. Statistical significance is highlighted with different alphabets.

Hm-oscar suppresses the masculinizing functions of lepidopteran masc genes

To confirm whether Hm-oscar suppresses the functions of the primary male sex determinant masc, their interactions were assessed through transfection in BmN-4 cells. These cells, derived from B. mori ovaries (Grace, 1967), exhibit female-type default sex determination (Kiuchi et al., 2014). However, overexpression of the masc gene from various lepidopteran insects, such as O. furnacalis, induces male-type sex determination by promoting the expression of BmdsxM and BmIMPM in BmN-4 cells (Fukui et al., 2015; Katsuma et al., 2019; Katsuma et al., 2022). Since BmIMPM regulates the splicing of BmdsxM and is exclusively expressed in males, we selected BmIMPM as a quantitative marker of masculinization, as described in Katsuma et al. (2022).

In contrast with the non-expressed control groups, which exhibited default female-type sex determination, cells expressing Hmmasc exhibited male-type sex determination (Fig. 4a). This was indicated by the increased expression of the male-specific splicing variant BmIMPM, although this increase was not statistically significant. Furthermore, the masculinizing function of Hmmasc was suppressed when Hmmasc and Hm-oscar were co-expressed, as evidenced by the low expression levels of BmImpM (Fig. 4a). wFur-encoded Oscar, which differs from Hm-Oscar in structure and amino acid sequence (Fig. 1a), also suppressed Hmmasc, but its function was less active than that of Hm-oscar (Fig. 4a). Hm-oscar also suppressed the function of masc genes derived from various lepidopteran insects (i.e., O. furnacalis, Spodoptera frugiperda, Bombyx mori, Trilocha varians, and Papilio machaon), which are highly divergent in their amino acid sequences (Fig. 4b,c). These results suggest that Hm-oscar has a broad spectrum of activity in Lepidoptera.

Hm-oscar suppressed the masculinizing function of lepidopteran Masc

(a) Relative expression levels of the male-specific BmImpM variant in transfected BmN-4 cells. The relative expression of BmImpM in the control (non-inserted pIZ/V5) and Hm-oscar/oscar co-transfected groups was normalized by setting the mean in the Hmmasc single transfected group as 100. (b) Relative expression levels of the BmImpM in Hm-oscar/masc transfected BmN-4 cells. The relative expression of BmImpM in each condition was normalized by setting the mean in the Hmmasc single transfected group as 100. The dot plots show all data points individually. Each experimental condition was replicated three times. Abbreviations: Hm, Homona magnanima; Bm, Bombyx mori; Of, Ostrinia furnacalis; Pm, Papilio machaon; Sf, Spodoptera frugiperda; and Tv, Trilocha varians. (c) Phylogeny and homologies of the Masc protein sequences used in this study. The sequence homology between each pair is presented in the figure and visualized as a heatmap, with red indicating high homology (Max = 100) and blue indicating low homology (Min = 23.82).

Discussion

In this study, we showed that the phage-encoded Hm-oscar, but not wmk, induced male lethality and a female-biased sex ratio in H. magnanima. Furthermore, the overexpressed Hm-oscar impaired male sex determination in Homona, recapitulating the wHm-t-induced phenotypes. Cell-based assays confirmed that Hm-oscar suppressed the masculinizing functions of masc. These results strongly suggested that Hm-oscar underlies the wHm-t-induced MK function in H. magnanima. In light of our results and previous findings in Homona, Ostrinia, and Bombyx (Arai et al., 2023a; Katsuma et al., 2022; Kiuchi et al., 2014), we hypothesize the following molecular mechanisms for wHm-t-induced MK: The Hm-Oscar protein targets and inhibits the function of the HmMasc protein, thereby resulting in female sex determination and disrupting dosage compensation, which ultimately leads to male death. Although the full mechanisms remain unclear, unbalanced Z-linked gene expression mediated by improper dosage compensation—resulting in Z-linked gene expression levels twice as high as those in normal males—appear to be lethal for lepidopteran males (Kiuchi et al., 2014; Fukui et al., 2015; Visser et al., 2021).

The Wolbachia-encoded oscar homologs identified so far show diversity in sequence length and are classified as type I, II, and III based on their amino acid sequence homologies. Type I and II Oscar proteins, derived from MK-inducing Wolbachia, contain many ankyrin repeats and a proteinase domain (Arai et al., 2024a). Although long ankyrin repeat sequences are postulated to be critical for the function of wFur-encoded Oscar protein (1830 aa, type I) (Katsuma et al., 2022), our study revealed that the functions of Hm-Oscar protein (1181 aa, type II), which encodes fewer ankyrin repeats, were comparable with those of Oscar protein carried by wFur. Interestingly, type II oscar gene is also present in the feminizing Wolbachia wFem in Eurema butterflies (Arai et al., 2024a). oscar homologs, which inhibit the masculinizing function and induce female sex determination, may have a conserved function in Wolbachia-induced MK and feminization in Lepidoptera.

In contrast to the results of this study, we have previously demonstrated that the phage-encoded wmk, but not Hm-oscar, induces male lethality in D. melanogaster (Arai et al., 2023b). Although the means of expression are different (i.e., transient in Homona and transgenic in Drosophila), this finding highlighted the differences in the mode/range of action of Wolbachia genes between insect species. It has been hypothesized that microbes induce MK in insects by targeting molecular mechanisms involved in sex determination and differentiation (Hornett et al., 2022). Sex determination systems in insects are diverse; for example, Lepidoptera (including H. magnanima) and Diptera (including D. melanogaster) do not share any known sex-determining genes other than dsx (Suzuki, 2018). The different outcomes in Homona and Drosophila are probably due to their different sex determination systems. Because oscar interacts with and suppresses masc, Hm-oscar could induce mortality in Homona males that possess masc, but not in Drosophila males that lack it. Considering that oscar homologs are not present in known MK Wolbachia strains in dipteran insects (Arai et al., 2024c; Katsuma et al., 2022), the mechanisms (i.e., causative genes) of Wolbachia-induced MK probably differ between insect taxa (e.g., between Lepidoptera and Diptera). While the mechanisms underlying wmk-induced lethality remain unclear, the distinct effects associated with this gene between Homona and Drosophila may also reflect their genetic backgrounds, such as the presence or absence of host factor(s) that interact with wmk. In addition, Katsuma et al. (2022) reported that the wmk homologs encoded by wFur did not affect the masculinizing function of masc in vitro, indicating that wmk likely targets factors other than masc. Whilst we cannot rule out the possibility that wmk may work synergistically or interactively with oscar in vivo— potentially acting downstream of oscar’s impact—our results strongly suggested that Wolbachia strains have acquired multiple MK genes through evolution.

Wolbachia-induced phenotypes are known to be influenced by the genetic backgrounds of hosts (Hornett et al., 2006; Sasaki et al., 2002; Veneti et al., 2012). Our study showed that the wHm-t-encoded Hm-oscar suppresses the function of masc in H. magnanima more efficiently than the wFur-encoded oscar, suggesting that this Wolbachia factor has undergone evolutionary tuning to adapt to its natural host. However, the mere presence of oscar and wmk homologs does not ensure MK expression. For example, wCauA, bearing type I oscar, did not induce MK in Ephestia (Cadra) cautella collected around 2000, although it did induce MK when transferred to the closely related host Ephestia kuehniella (Sasaki et al., 2002). An MK phenotype, presumably induced by wCauA, was also recorded in E. cautella around the 1970s (Takahashi & Kuwahara, 1970). These findings suggest a suppressor evolution against wCauA-induced MK in E. cautella. In addition, Wolbachia typically encode a number of wmk homologs in their genomes, but their effects differ among insects and are not apparent in some species such as Homona, possibly due to the diversity in sex determination system or the evolution of suppressors. Nonetheless, given the high copy number of wmk homologs in Wolbachia genomes, they may still contribute to Wolbachia biology as transcriptional regulators and potential virulence factors. Virulence genes often undergo duplication and substitution under strong selective pressure (Hill et al., 2022; Jones & Dangl, 2006). An intense evolutionary arms race between Wolbachia and their hosts could have increased and diversified the MK-associated Wolbachia genes. Conversely, natural selection favors the rescue of males by suppressing the Wolbachia-induced reproductive manipulations (Hornett et al., 2006; Hornett et al., 2014), which may involve changes in the sex determination system because Wolbachia strains frequently hijack host reproduction systems. In this context, Wolbachia and other MK-inducing microbes may be a hidden driver for the diversification of complex insect sex determination systems.

In this study, we clarified the conserved roles of the Wolbachia-encoded oscar homologs in Lepidoptera and demonstrated that Wolbachia have evolved distinct MK mechanisms (through causative genes) in insect taxa. The diversification of phenotype/virulence-associated genes and the rampant horizontal transmission of phages carrying virulence genes between Wolbachia strains may have contributed to the outstanding success of this bacterial genus. In addition to oscar and wmk, Wolbachia may retain other uncharacterized genes that induce male lethality, and further studies on diverse Wolbachia–host systems are highly warranted. Our findings provide insights into the molecular mechanisms and evolutionary relationships between endosymbionts and their hosts, which may also contribute to the design of pest management strategies.

Materials and methods

Experimental model and subject details

Homona magnanima

A laboratory-maintained H. magnanima line with a normal sex ratio (NSR) that was negative for Wolbachia and other endosymbionts, was used in our experiments. This line was initially collected in Tokyo, Japan, in 1999 and has been maintained inbred for over 250 generations in the laboratory. Larvae were reared using the artificial SilkMate 2S diet (Nosan Co., Yokohama, Japan) at 25°C under a long photoperiod (16L:8D) (Arai et al., 2019).

The laboratory-maintained all female H. magnanima WT12 line, which was initially collected in Taiwan (Tea Research Extension Station, Taoyuan city) in 2015 (Arai et al., 2020), was also used in this study. This line was maintained for over 50 generations by crossing it with the males of the NSR line.

BmN-4 cell line

BmN-4 cells (provided by Chisa Yasunaga-Aoki, Kyushu University, and maintained in our laboratory) were cultured at 26°C in IPL-41 Insect Medium (Applichem, Darmstadt, Germany) supplemented with 10% fetal bovine serum.

Transient expression of MK-associated phage genes

mRNA synthesis

Codon-optimized wmk genes (wmk-1 to wmk-4), conjugated wmk pairs using a T2A peptide (i.e., wmk-1-T2A-wmk-2 and wmk-3-T2A-wmk-4), and Hm-oscar genes synthesized by Arai et al. (2023b) were used for mRNA synthesis. These synthetic genes were ligated into the plasmid pIZ/V5-His (Invitrogen, MA, USA) using the NEBuilder® HiFi DNA Assembly kit (New England Biolabs, MA, USA) following the manufacturer’s protocol. The inserts of the vector (i.e., wmk-1, wmk-2, wmk-3, wmk-4, Hm-oscar, wmk-1-T2A-wmk-2, wmk-3-T2A-wmk-4, and GFP as a control gene) were amplified using KOD-one (TOYOBO, Osaka, Japan) with the primer set containing the T7 promoter described in Fukui et al. (2015) (i.e., pIZ-F-T7: 5′-TAATACGACTCACTATAGGGAGACAGTTGAACAGCATCTGTTC-3′ and pIZ-R: 5′-GACAATACAAACTAAGATTTAGTCAG-3′) under the following PCR conditions: 20 cycles at 98°C for 10 s, 62°C for 5 s, and 68°C for 15 s. The amplicons were purified using the QIAquick PCR purification kit (Qiagen, Hilden, Germany), and 100–200 ng DNA was used for mRNA synthesis. The capped mRNA (cRNA) with poly(A) tail was synthesized using the mMESSAGE mMACHINE® T7 Ultra kit (Invitrogen, MA, USA) following the manufacturer’s protocol with some modifications. In brief, the assembled transcription reaction (10 µL of T7 2X NTP/ARCA, 2 µL of 10X T7 reaction buffer, 2 µL of T7 enzyme mix, and 6 µL of linear DNA template diluted with nuclease-free water) was incubated at 37°C for at least 10 h to maximize RNA yields. After poly(A) tailing, cRNA was purified using ISOGEN II (Nippongene, Tokyo, Japan) and dissolved in up to 10 µL of nuclease-free water to achieve an RNA concentration of approximately 1500–2000 ng/µL. The synthesized cRNA was preserved at −80°C until further use.

Preparation of H. magnanima eggs

A total of 15 males and 10 females of H. magnanima were mated in a plastic box (30 cm × 20 cm × 5 cm) for 3–4 days. The collection of egg masses began the day after oviposition was confirmed. In brief, newly oviposited egg masses were collected at 30-min intervals during the dark period using red light (which the moths cannot perceive). Females started to oviposit eggs at least 4 h into the dark period and, within less than 30 min post oviposition (mpo), the egg masses were collected. The collection lasted until the start of the light period. The egg masses were then subjected to microinjection assays.

Inoculum preparation, microinjection, and maintenance of the injected embryos

A glass needle for microinjection was prepared using glass capillary GD-1 (Narishige, Tokyo, Japan) with a PC-10 puller (Narishige). The glass capillaries were pulled at two temperatures (first stage: 75°C, second stage: 65°C) using two heavy weights and one light weight. The movement position during the second heating stage was set to 3 mm (range 1–10 mm).

The cRNA solution was diluted in a buffer (100 mM potassium acetate, 2 mM magnesium acetate, and 30 mM HEPES-KOH; at pH 7.4) containing 0.2% (W/V) Brilliant Blue FCF (Wako, Osaka, Japan) to obtain an RNA concentration of 1000 ng/µL. Approximately 1–4 µL of dye-containing mRNA solution was aspirated into the glass needles (capillaries), and the edge of each needle was ground using Micro Grinder EG-402 (Narishige) at an angle of 20 degrees for 2 s.

The fresh egg masses of H. magnanima (collected at less than 30 mpo as described above) were put on a double-sided sticky tape (15 mm × 5 m T4612, Nitoms, Tokyo, Japan) and placed on glass slides. Under a Nikon SMZ1270 microscope (Nicon, Tokyo, Japan), the RNA solution (30–100 fL) was injected into individual eggs using an microinjector IM-400 (Narishige) (balance pressure set at 0.010–0.030; injection pressure set at 0.050-0.100) and a QP-3JOY-2R electric micromanipulator (MicroSupport, Shizuoka, Shizuoka).

The injected egg masses were maintained in a 90-mm plastic Petri dish fitted with a slightly wet filter paper. As high humidity interfered with the development of the embryos, the injected eggs (egg mass) were first maintained at a relative humidity of 30%–50% (0–3 days post injection) and then at a higher humidity until hatching (4–6 dpo, 60% relative humidity). The hatched larvae were reared separately with 1/2 ounces of SilkMate 2S (Nosan Co.) until eclosion. The adult H. magnanima moths that emerged from the cRNA-injected embryos were sexed based on their external morphology.

Sexing of embryos/neonates and RNA extraction

To verify the effect of the transient expression of Hm-oscar on sex determination in H. magnanima, Hm-oscar/GFP-expressed, non-injected, and wHm-t-infected mature embryos showing black head capsule (1 day before hatching, 5–6 days post oviposition, dpo) were dissected on glass slides using forceps, as described in Arai et al. (2022). Their Malpighian tubules were fixed with methanol/acetic acid (50% v/v) and stained with lactic acetic orcein for W chromatin observations. The remaining tissues not used for sexing were stored in ISOGEN II (Nippon Gene) at −80°C until subsequent extraction. In total, 12 males or females (confirmed based on the presence or absence of W chromatin) were pooled and homogenized in ISOGEN II to extract RNA as described in Arai et al. (2023a). In brief, 0.4 times the volume of UltraPure distilled water (Invitrogen) was added to the ISOGEN II homogenates, which were centrifuged at 12000 × g for 15 min at 4°C to pellet proteins and DNAs. The resulting supernatant was mixed with the same volume of isopropanol to precipitate RNAs; then, the resulting solutions were transferred to EconoSpin columns (Epoch Life Science) and centrifuged at 17,900 × g and 4°C for 2 min. The RNAs captured in the column were washed twice with 80% ethanol and eluted in 20 µL of UltraPure distilled water (Invitrogen).

Hmdsx detection

Sex-specific dsx splicing variants of H. magnanima were detected as described in Arai et al. (2023a). In brief, total RNA (100–300 ng) extracted from sex-determined mature embryos was reverse-transcribed using PrimeScript− II Reverse Transcriptase (TaKaRa, Shiga, Japan) at 30°C for 10 min, 45°C for 60 min, and 70°C for 15 min. Then, cDNA was amplified using KOD-FX Neo (Toyobo Co., Ltd.) with the following two primers: Hmdsx_long3F (5′-TGCCTAAAGTGAAAACGCCGAGGAGCC-3′) and Hmdsx_Mrev (5′-TGGAGGTCTCTTTTCATCCGG-3′). The PCR conditions used were as follows: 94°C for 2 min, followed by 45 cycles of 98°C for 10 s, 66°C for 30 s, and 68°C for 30 s. The amplicons were subjected to electrophoresis on 2.0% agarose Tris-borate-EDTA buffer (89 mM Tris-borate, 89 mM boric acid, 2 mM EDTA) gels.

Quantification of Z chromosome-linked genes

The effects on dosage compensation were assessed by measuring differences in gene expression, as described in Fukui et al. (2015) and Arai et al. (2023a). A total of 1.0 µg of the total RNA extracted from Hm-oscar or GFP-overexpressed mature embryos (108 hpo, approximately 40-50 sex determined embryos) was used to prepare mRNA-seq libraries via the NEBNext Poly (A) mRNA Magnetic Isolation Module (New England Biolabs) and the NEBNext Ultra II RNA Library Prep kit for Illumina (New England Biolabs) following the manufacturer’s protocol. The adaptor sequences and low-quality reads (Qscore <20) were removed from the generated sequence data [150 bp paired-end (PE150)] using Trimmomatic (Bolger et al., 2014). The trimmed reads were mapped to the previously assembled transcriptome database for H. magnanima (Arai et al., 2023a) using Kallisto (Bray et al., 2016) to generate the normalized read count data (transcripts per million, TPM). While we did not generate new RNA-seq data of the wHm-t-infected and uninfected groups, the expression data included in Arai et al. (2023a) were re-analyzed along with the newly obtained RNA-seq data above. The binary logarithms of the TPM differences between males and females belonging to each H. magnanima group (i.e., Hm-oscar/GFP-expressed) were calculated to assess the fold-changes in gene expression levels between sexes. As described in Arai et al. (2023a), the transcriptome data of H. magnanima were annotated using the B. mori gene sets obtained from KAIKOBASE (https://kaikobase.dna.affrc.go.jp). The binary logarithms of the TPM differences between males and females in the B. mori chromosomes 1 to 28 were plotted.

To confirm the effects observed in the RNA-seq-based quantification of the Z chromosome genes, expression level of the Z-linked Hmtpi gene was quantified by reverse transcription-qPCR as described in Arai et al. (2023a). Briefly, total RNA (100– 300 ng) extracted from pooled male or female mature embryos (108 hpo) was reverse transcribed using PrimeScript II reverse transcriptase (TaKaRa) as described above. The cDNA was used to quantify relative gene expression levels, with normalization to the control gene elongation factor 1a (ef1a). qPCR was performed using primer sets for Hmtpi (HmTpi_F180702_212: 5′-GCTGCGAGTGGGTGATTTTG-3′; HmTpi_R180702_326: 5′-GCGATCACTTTCAGACCCGA-3′) and Hmef1a (Hmef1a_F_val1_85: 5′-TTTCCAGGGTGGTTGAGCA-3′; Hmef1a_R_val1_193: 5′-CCGTTAAGGAGCTGCGTCG-3′), and KOD SYBR qPCR Mix (Toyobo, Osaka, Japan) in a LightCycler 96 system (Roche, Basel, Switzerland). The qPCR reaction was made with 5 µl KOD SYBR (TOYOBO), 0.4 µL forward primer [10 pmol/µl], 0.4 µl reverse primer [10 pmol/µl], and 3.2 µl water. The qPCR was performed under the following conditions: 180 s at 95°C, 40 cycles of 8 s at 98°C, 10 s at 60°C, and 10 s at 68°C, followed by heating to 90°C for melting curve analysis. Mean cycle threshold (CT) values of samples were calculated for at least 13 replicates, and both ΔCT (CT Z gene – CT ef1a) and ΔΔCT (ΔCTmale – ΔCTfem) values were calculated.

Transfection assays and quantification of BmIMPM

The coding sequence of Hmmasc was amplified from cDNA derived from the RNA extracted from male embryos of the NSR line using KOD-one (TOYOBO) with the following primer set: HmMasc_CDS_HindIIIF: 5′-GCAAAGCTTCAACATGATCTCTCGCCAACCACAATCAACATCA-3′ and HmMasc_CDS_BamHI R: 5′-GCAGGATCCCAACCTACTGATAAGGAGGGAAGTAAGGCTGCTG-3′. The following PCR conditions were applied: 20 cycles of 98°C for 10 s, 62°C for 5 s, and 68°C for 15 s. The amplicon was purified with the QIAquick PCR purification kit (Qiagen) and cloned into the pIZ/V5-His vector using the HindIII and BamHI enzymes (New England Biolabs). The codon-optimized oscar gene of wFur (Katsuma et al. 2022) was also cloned into the pIZ/V5-His vector using the KpnI and NotI enzymes (TaKaRa).

To verify the masculinizing function of Hmmasc, BmN-4 cells (4 × 105 cells per dish, diameter 35 mm) were transfected with 1 µg of each plasmid DNA (pIZ/V5-His having Hmmasc) using FuGENE HD (Promega, WI, USA), as described in Katsuma et al. (2022). To clarify whether Hm-oscar suppressed the function of Hmmasc, 1 µg of plasmid DNA (pIZ/V5-His bearing either Hmmasc or Hm-oscar and non-inserted vector as negative control) was transfected to the BmN-4 cells using FuGENE HD (Promega). Three days after transfection, the cells were collected and subjected to RNA extraction via TRI REAGENT® (Molecular Research Center Inc., USA) and cDNA construction with AMV transcriptase (TaKaRa). The degree of masculinization in the BmN-4 cells (default female-type sex determination) was verified by quantifying the expression levels of BmIMPM (Suzuki et al., 2010), which is involved in male-specific sex determination cascades, using primers BmIMP_F: 5′-ATGCGGGAAGAAGGTTTTATG-3′ and BmIMP_R: 5′-TCATCCCGCCTCAGACGATTG-3′, as described in Fukui et al. (2015). BmRp49 gene was also amplified as a control gene using primers rp49_F: 5′-CCCAACATTGGTTACGGTTC-3′ and rp49_R: 5′-GCTCTTTCCACGATCAGCTT-3′. Further interactions between Hm-oscar and masc genes derived from lepidopteran insects [i.e., Trilocha varians masc (Tvmasc), Spodoptera frugiperda masc (Sfmasc), B. mori masc (Bmmasc), O. furnacalis masc (Ofmasc), and Papilio machaon masc (Pmmasc)] (Katsuma et al., 2022b) were assessed using the same procedures. The expression levels of BmIMPM were analyzed as described in Katsuma et al. (2022). In brief, the relative expression of BmIMPM (CT BmIMPM / CT BmRp49) was calculated for each experimental group. Then, relative expression of BmIMPM of each sample was normalized by setting the mean in the Hmmasc singly transfected group as 100. The normalized relative expression of BmIMPM was visualized using R v4.0 with ggplot2 (https://ggplot2.tidyverse.org/).

Quantification and statistical analysis

The number of surviving H. magnanima injected with either Hm-oscar, wmks, or GFP in each replicate were counted at the adult stage. The male ratios (number of adult males/numbers of all adults) under all conditions were compared using the Steel– Dwass test in R v4.0. The sex ratio bias was also assessed based on the total numbers of (i) male and female adults and (ii) male and female embryos under each condition using the binomial test in R v4.0.

Gene expression levels between chromosomes (RNA-seq), a Z-linked Hmtpi gene expression in H. magnanima (RT-qPCR), and normalized BmIMPM expression in BmN-4 cells under all conditions (RT-qPCR) were compared using the Steel–Dwass test in R v4.0.

A phylogenetic tree of Masc proteins used in this study was constructed based on maximum likelihood with bootstrap re-sampling of 1,000 replicates using MEGA7 (Kumar et al., 2016).